\chapter{ Inputs-outputs}

\section{ Preliminary remark}

During a computation, the \tomawac software uses a number of files, some of which are
optional, as inputs and outputs.

 The input files are:

\begin{itemize}
\item  The steering or CAS file (mandatory),

\item  The mesh or geometry file (mandatory),

\item  The boundary conditions or CONLIM file (mandatory),

\item  The seabed, bottom or bathymetry file (optional),

\item  The FORTRAN or PRINCI file (optional),

\item  The currents file (optional),

\item  The winds file (optional),

\item The imposed spectra on the boundary (optional),

\item The coordinates of the nodes for which a spectra will be imposed (optional),

\item  The previous computation file (optional),

\item  The binary user file (optional),

\item  The formatted user file (optional),

\item The coordinates of the nodes for which spectra will be written (mandatory if a
  punctual or a spectra file is defined),

\item The time series coordinates file which contains coordinates and periods of
  time where time series are extracted and written in the time series file.
\end{itemize}

The output files are:

\begin{itemize}
\item  The 2D results or grid file (mandatory),
\item  The punctual results or spectra file (optional),
\item  The next computation file (optional),
\item  The listing printout (either on the display screen or in the file, see in
  \ref{se:usefulsub}),
\item  The binary user file (optional),
\item  The formatted user file (optional),
\item  The time series file which contains time series at locations defined in
  the time series coordinates file.
\end{itemize}

\section{ The files}

\subsection{ The steering (or CAS) file}
\label{se:steeringfile}
The steering file name is specified in the steering file through the keyword:
\textit{STEERING FILE.}

It is a text file created by means of a text editor. In a way, it serves as the
computation control panel. It includes a set of keywords to which values are assigned.
If a keyword does not appear in this file, then \tomawac will assign to it the default
value as defined in the dictionary file (refer to the reference manual). If such a
default value is not defined in the dictionary, then the computation will come to a
halt and display an error message. For instance, the command \textit{NUMBER OF
  DIRECTIONS} = 12 is for specifying that the direction spectrum of wave action or its
moments will be discretised over 12 propagation directions.

 \tomawac reads the steering file at the beginning of the computation.

 Both dictionary file and steering file are read by the so-called DAMOCLES utility which
 is included in \tomawac. The syntactic rules of DAMOCLES should then be observed upon
 the creation of the steering file These rules are described here below.

 The write rules are as follows:

\begin{itemize}
\item  The keywords can be of the Integer, Real, Logical or Character format type.

\item  The keyword sequence order in the steering file is of no importance.

\item  Each line has a maximum of 72 characters. However, as many linefeeds as one wants
  are allowed provided that the keyword name does not run from one line to the next.

\item  For the table-like keywords, the successive values are separated by a semi-colon.
  A number of values equal to the table dimension should not necessarily be given; in
  such a case, DAMOCLES returns the number of values being read. For example:
\end{itemize}

\textit{ABSCISSAE OF SPECTRUM PRINTOUT POINTS }= 1.2;3.4   (that keyword is declared
as a 19-valued table)

\begin{itemize}
\item  The symbols ":" or "=" are indiscriminately used to separate a keyword from its
  value. They can be either preceded or followed with any number of blanks. The value
  itself may appear on the next line. For example:    

\textit{NUMBER OF DIRECTIONS} = 12\\
or \\   \textit{NUMBER OF DIRECTIONS}: 12 \\  or else \\  \textit{NUMBER OF
  DIRECTIONS} =\\   12

\item  The characters occurring between a pair of "/" on one line are regarded as
  comments. Likewise, the characters occurring between a "/" and a the end of a line
  are regarded as comments. For example:  \textit{TYPE OF BOUNDARY DIRECTIONAL
    SPECTRUM }= 1 / Jonswap spectrum

\item  A whole line beginning with a "/" in the first column is regarded as a comment,
  even though another / occurs on the line. For example:  / The geometry file is .
  /maillage/geo

\item  Integer writing: Do not exceed the maximum size being allowed by the machine
  (in a machine with 32 bit architecture, the values range from -2 147 483 647 to
  +2 147 483 648. Do not enter a blank between the sign (optional for the + sign) and
  the number. A dot at the end of the number is tolerated.

\item  Real writing: A dot or a comma is allowed as a decimal point, as well as the
  FORTRAN E and D formats (1.E-3 0.001 0,001 1.D-3 denote the same value).

\item  Logical value writing: The values 1, YES, OUI, .TRUE., TRUE, VRAI on the one
  hand, and 0, NON, NO, .FALSE., FALSE, FAUX on the other hand are allowed.

\item  Character string writing: Those strings including blanks or reserved symbols
  ("/",":", "=", "\&") should be put in single quotes ('). The value of a character
  keyword may include up to 144 characters. As in FORTRAN, the quotes occurring within
  a string should be doubled. A string may neither begin nor end with a blank.
  For example:   \\
\textit{TITLE} = 'HOULE D''OUEST'
\end{itemize}

In addition to the keywords, a number of directives or metacommands that are
interpreted during the sequential readout of the steering file may be used as well:

\begin{itemize}
\item  The \textit{\&FIN} command indicates the end of file (even though the file is
  not completed). Thus, some keywords can be disabled simply by placing them behind
  that command for easily making it possible to enable them again subsequently.

\item  The \textit{\&ETA} command prints the list of keywords and the relevant values
  at the time when DAMOCLES meets that command. This display will occur at the beginning
  of listing printout.

\item  The \textit{\&LIS} command prints the list of keywords. This display will occur
  at the beginning of listing printout.

\item  The \textit{\&IND} command prints the detailed list of keywords. This display
  will occur at the beginning of listing printout.

\item  The \textit{\&STO} command causes the interruption of the program, the
  computation does not go on.
\end{itemize}


\subsection{ The geometry file}

The geometry file name is specified in the steering file through the keyword:
\textit{GEOMETRY FILE}.

It is either a SERAFIN-formatted or a MED-formatted binary file: it can be read by
FUDAA PRE-PRO or RUBENS (for SERAFIN) or Salome (for med) and it can be created by the
STBTEL module from the file(s) as produced by the mesh generator. The format structure
is described in \ref{se:filestruct}

This file includes the complete information about the horizontal mesh, i.e. the number
of mesh points (variable NPOIN2), the number of elements (variable NELEM2), the X and Y
vectors containing the co-ordinates of all the points and, lastly, the IKLE2 vector
containing the connectivity table.

Furthermore, this file may also include bathymetry information in each point of the
mesh, provided that the interpolation of the bathymetry was carried out during the
execution of the STBTEL module or during the generation of the mesh.

\tomawac reproduces the information regarding the geometry at the beginning of the 2D
results. Any computation results file can then be used as a geometry file when one wants
to perform a further simulation on the same mesh.


\subsection{ The boundary conditions file. }

The boundary conditions file name is specified in the steering file through the
keyword: 

\textit{BOUNDARY CONDITIONS FILE.}

It is a formatted file that can be created automatically by STBTEL and can be modified
by means of a text editor.
If the geometry file is a serafin, Each line in this file is assigned to one point of
the boundary and listed in sequential order in terms of the boundary node numbers. The
numbering of the boundary points first delineates the domain contour in the
counterclockwise direction, then the islands in the clockwise direction. This file is
described in detail in \ref{se:BCfile}

If the geometry file format is med, then each line correspond to a group of segments
defined in the geometry file.

\subsection{ The currents file}
\label{se:currentfile}
According to its type -- binary or formatted- the currents file name is specified in
the steering file through the keywords: \textit{BINARY CURRENTS FILE} and\textit{
  FORMATTED CURRENTS FILE.}

It is the file from which \tomawac reads the current field components. The current
field may be either stationary or non-stationary. The current field will be
non-stationary when the keyword \textit{CONSIDERATION OF TIDE }is set to TRUE. When
the current field is stationary, the keyword \textit{CONSIDERATION OF A STATIONARY
  CURRENT} should be set to TRUE. By default, both keywords will be set to FALSE.
When both are set to TRUE, the keywords will be inconsistent, and the program will halt.

Several commonly used formats can be read. This selection is made through the integer
keyword \textit{CURRENTS FILE FORMAT}. It can be set to a value from 3 to 4.

\begin{itemize}
\item  The format is 3: it is a TELEMAC result file of the SERAFIN or MED standard.
  It is a binary file the name and its name be assigned to the keyword: \textit{BINARY
    CURRENTS FILE}. If the current is assumed to be stationary, then the additional
  keyword \textit{TIME INCREMENT NUMBER IN TELEMAC FILE }should be used in order to
  find the time step number related to the desired record. TELEMAC data other than
  the current components e.g. water levels, can also be read by means of this format
  (refer to \ref{se:telemacdata}).

\item  The format is 4: data written in a different format can be read provided that
  the user supplies the relevant subroutine in the relevant FORTRAN file (see
  \ref{se:current} and \ref{se:tide}).
\end{itemize}


\subsection{ The tidal water level file}
\label{se:tidalfile}
According to its type -- binary or formatted- the tidal water level file name is
specified in the steering file through the keywords: \textit{BINARY TIDAL WATER LEVEL
  FILE }or\textit{ FORMATTED TIDAL WATER LEVEL FILE.}

This is the file from which \tomawac reads the tidal water level being referred to
the \textit{INITIAL STILL WATER LEVEL}. Several commonly used formats can be read.
This selection is made by means of the integer keyword \textit{TIDAL WATER LEVEL FILE
  FORMAT}. It can be set to a value from 3 to 4.

\begin{itemize}
\item  The format is 3: it is a TELEMACTELEMAC result file of the SERAFINSERAFIN
  standard. It is a binary file and its name should be assigned to the keyword:
  \textit{BINARY TIDAL WATER LEVEL FILE.}

\item  The format is 4: data written in a different format can be read provided that
  the user supplies the relevant subroutine in the relevant FORTRAN file (see in
  \ref{se:tide}).
\end{itemize}


\subsection{ The winds file}
\label{se:windfile}

According to its type -- binary or formatted- the wind file name is specified in the
steering file through the keywords: \textit{BINARY WINDS FILE }or\textit{ FORMATTED
  WINDS FILE.}

This is the file from which \tomawac reads the information about the wind fields. As
in the case of the current, several read formats are allowed. The integer keyword
\textit{WINDS FILE FORMAT} can be set to values from 1 to 4.

\begin{itemize}
\item  The format is 3: it is a TELEMAC result file of the SERAFIN standard. It is a
  binary file and its name should be assigned to the keyword: \textit{BINARY WINDS
    FILE}. If the wind is assumed to be stationary, then the additional keyword
  \textit{TIME STEPTIME STEP NUMBER IN TELEMAC FILE }should be used in order to find
  the time step number related to the desired record.
\item  The format is 4: data written in a different format can be read provided that
  the user supplies the relevant subroutine in the relevant FORTRAN file (see in
  \ref{se:wind}).
\end{itemize}

\subsection{ The imposed spectra file and the file containing the associated coordinates}
\label{se:ImpSpeCoorFiles}

It is possible to impose spectra on the open boundary of the mesh. To do so, the
boundary condition for those nodes need to follow the format \textit{5 4 4 4}, see
section \ref{se:BCfile}. There are then two files that need to be 
given:

\begin{itemize}
\item \textit{IMPOSED SPECTRA FILE}: which is a binary file, such as outputted in the
  punctual or spectrum
results file, see section \ref{se:SpeFile}.

\item \textit{FILE WITH COORDINATES OF SPECTRA TO IMPOSE}: which is an ascii file with
  the coordinates associated
with each spectrum in the \textit{IMPOSED SPECTRA FILE}.
\end{itemize}

Since the \textit{IMPOSED SPECTRA FILE} follows the format of the \telemacsystem{}
results, the format can be given using
the keyword \textit{IMPOSED SPECTRA FILE FORMAT}. In addition, since imposed spectra
can change in time, the
\textit{TIME UNIT OF IMPOSED SPECTRA FILE} and \textit{TIME SHIFT OF IMPOSED SPECTRA
  FILE} can be modified.

Finally, the \textit{FILE WITH COORDINATES OF SPECTRA TO IMPOSE} follows the format of
DAT outputs of the module {\scshape SMESH} in \salome{}. For a list of nodes, it follows
this structure:

\lstset{language=TelemacCas,
        basicstyle=\scriptsize\ttfamily}
\begin{lstlisting}[frame=trBL]
<Number of Points> 0
<Point Id> <X> <Y> <Z>
...
\end{lstlisting}

\subsection{The area definition file.}
When a user is using vegetation or porous medi he has the opportunity to define areas
that will contains vegetation or porous media through a list of polygons. Those
polygones are defined through the {\it FILE WITH DEFINITION OF POLYGONS}. This file is
an i2s format file. The first line contains the number of polygons. The second one, the
numbers of points of each polygon and then each line contains the coordinates of each
point. 


\subsection{ The previous computation file}

This previous computation file name is specified in the steering file through the
character keyword: \textit{PREVIOUS COMPUTATION FILE.}

If a \textit{NEXT COMPUTATION} is done\textit{, }\tomawac fetches this file in order
to initialize the directional spectrum of wave action at every point. This file's
format, which is specific to \tomawac, is described in \ref{se:desformat}. It is a
 binary file.

\subsection{ The global results file}

The global results file name is specified in the steering file through the keyword:
\textit{GLOBAL RESULTS FILE.}

This file is created when a \textit{GLOBAL OUTPUT AT THE END} is requested. It saves
the wave action density directional spectrum at every point in the last time step. This
file format is described in \ref{se:desformat}.


\subsection{ The 2D results file }

The 2D results file name is specified in the steering file through the character
keyword: \textit{2D RESULTS FILE.}

This is the file into which \tomawac writes the results of the 2-dimensional variables
during the computation. It is a binary file of the \telemac standard (SERAFIN or MED).
The data contained in it are in the following order:

\begin{itemize}
\item  all the data about the mesh geometry;
\item  the names of the variables being stored;
\item  for each time step, the time and the values of the variables are given for each
  point of the 2D mesh.
\end{itemize}

 Its content varies according to the values of the following keywords:

\begin{itemize}
\item  \textit{NUMBER OF FIRST ITERATION FOR GRAPHICS PRINTOUTS:} provided for
  determining from which time step will the data storage desirably begin, so that
  the file size will not be too large.
\item  \textit{PERIOD FOR GRAPHICS PRINTOUTS}: sets the period, as a number of
  propagation time increments, of printouts so that the file size will not be too large.
\item  \textit{VARIABLES FOR 2D GRAPHICS PRINTOUTS}: provided for specifying the
  list of variables to be stored into the 2D results file. Each variable is identified
  by 2, 3 or 4 letters (refer to \ref{tab:resultvar} that lists the available variables).
\end{itemize}

\begin{longtable}{rl}
%%%%%%%%%%%%%%
% First Header
%%%%%%%%%%%%%%
\caption{List of 2D results variables}
\label{tab:resultvar}
\\
%
\toprule
\toprule
\textbf{Identifier} & \textbf{Definition} \\
\midrule
%
\endfirsthead
%%%%%%%%%%%%%%
% Other Headers
%%%%%%%%%%%%%%
\caption{(continued)}\\
%
\toprule
\toprule
\textbf{Identifier} & \textbf{Definition} \\
\midrule
%
\endhead
%%%%%%%%%%%%%%
% Foot
%%%%%%%%%%%%%%
\bottomrule
\bottomrule
%
\endfoot
%%%%%%%%%%%%%%
% Main table
%%%%%%%%%%%%%%
%
M0 &  Total variance \\
HM0  &  Spectral significant wave height  \\
DMOY  &  Mean wave direction  \\
SPD  &  Mean directional spreading  \\
ZF &  Sea bottom level  \\
WD &  Water depth  \\
UX &  Current along X  \\
UY &  Current along Y  \\
VX &  Wind along X  \\
VY &  Wind along Y  \\
FX &  Driving force along X  \\
FY &  Driving force along Y  \\
SXX  &  Radiation stress along xx  \\
SYY  &  Radiation stress along yy  \\
SXY  &  Radiation stress along xy  \\
UWB  &  Bottom celerity  \\
PRI  &  Private table  \\
FMOY  &  Mean frequency FMOY  \\
FM01  &  Mean frequency FM01  \\
FM02  &  Mean frequency FM02  \\
FPD  &  Discrete peak frequency  \\
FPR5  &  Peak frequency by Read method of order 5  \\
FPR8  &  Peak frequency by Read method of order 8  \\
US &  Surface friction velocity u*  \\
CD &  Surface drag coefficient CD  \\
Z0 &  Surface roughness length Z0  \\
WS &  Surface wave stress  \\
TMOY  &  Mean period Tmoy  \\
TM01  &  Mean period Tm01  \\
TM02  &  Mean period Tm02  \\
TPD  &  Discrete peak period  \\
TPR5  &  Peak period by Read method of order 5  \\
TPR8  &  Peak period by Read method of order 8  \\
POW  &  Wave power (per meter along wave crest)  \\
BETA &  Breaking waves rate \\
BETAWC & White Capping rate \\
SRE & Surface Roller Energy \\
DBR & Breaker Dissipation \\
DSR & Roller Dissipation \\
%
\end{longtable}


For instance, if the significant wave heights, the water depths and the average wave
propagation directions are desired,

 \textit{VARIABLES FOR 2D GRAPHICS PRINTOUTS }= HM0,WD,DMOY

 must be entered in the steering file.


\subsection{ The time series file}
The time series file is a file (in SERAFIN or MED format), in which time series
are written for certain points in the domain of the simulation.
This feature has been introduced since release 8.5.
In this way, one can obtain results with a high time frequency, without using
a large amount of disk space.
The time series file module interpolates data (using linear interpolation
even if using quasi-bubble or quadratic elements)
to the required output coordinates.
\begin{itemize}
\item \telkey{TIME SERIES FILE} The name of the output file to write.
The time and coordinates correspond to those given in
\telkey{TIME SERIES COORDINATES FILE},
\item \telkey{TIME SERIES FILE FORMAT} The file format of
\telkey{TIME SERIES FILE} (single or double precision SERAFIN, or MED,
default = SERAFIN),
\item \telkey{TIME SERIES COORDINATES FILE} The file which contains the output
times and output coordinates for \telkey{TIME SERIES FILE}.
The format is described next.
\end{itemize}

The \telkey{TIME SERIES COORDINATES FILE} has following format:
\begin {itemize}
\item On the first line, there are two numbers indicating:
\begin{itemize}
\item The number of time periods for which data is written to the
\telkey{TIME SERIES FILE},
\item The number of coordinates for which output is generated.
\end {itemize}
\item On the next lines, for each output period, there are three numbers
indicating:
\begin{itemize}
\item Start of the output period (in seconds since the start of the computation),
\item End of the output period (in seconds since the start of the computation),
\item Output time interval (in seconds).
\end {itemize}
\item For each coordinate, three numbers and a string, containing respectively:
\begin{itemize}
\item $x$-coordinate of the output point,
\item $y$-coordinate of the output point,
\item Unique ID of each station (in order to easily encounter the output
data in the output files).
This ID will be written to the IKLE array in the output file, which has a size
[NPOINTS $\times$ 1].
Note that the time series file consists of a list of points.
There is no mesh information in a time series file.
Therefore, the length of the IKLE array is the same as the number of points in
the time series file,
\item The name of the station (for better readability of the coordinate file).
\end{itemize}
\item Anything after the list of coordinates is considered comment and ignored
by \tel.
\end {itemize}

An example of a \telkey{TIME SERIES COORDINATES FILE} is given below.

\begin{verbatim}
2 3
0 100 10
500 1000 20
250  175     01 output_point_01
500  175     02 output_point_01
750  175     03 output_point_02

Description of the file
1st line: number of periods and number of points
2nd line: period 1: start time, end time and interval (in seconds)
3rd line: period 2: start time, end time and interval (in seconds)
4th-6th line: output points; x and y coordinates, unique ID, and station name
\end{verbatim}


\subsection{ The punctual or spectrum results file}
\label{se:SpeFile}

This file's name is specified in the steering file through the character keyword:
\textit{PUNCTUAL RESULTS FILE.}

This is the file into which the directional spectra of wave action at some previously
specified points are stored by \tomawac during the computation. To define those points,
a list of coordinates needs to be given, and the spectra of the closest node will be
outputted. This means that there is no spatial interpolation done. There are two methods
that can be used to define those points:

\begin{enumerate}

\item \textbf{Through keywords in the steering file:}

The $X$ and $Y$ coordinates are given through the following keywords:
\textit{ABSCISSAE OF SPECTRUM PRINTOUT POINTS} and \textit{ORDINATES OF SPECTRUM
  PRINTOUT POINTS}.
These are chart keywords, which means that each value is seperated with semicolon, and
the number of printout points is equal to the size of the array given. For example:

\begin{CommentBlock}{Keywords to write a spectra using method 1:}
\lstset{language=TelemacCas,
        basicstyle=\scriptsize\ttfamily}
\begin{lstlisting}[frame=trBL]
/--------------------------------------------------------------------/
/ WRITING SPECTRA
/--------------------------------------------------------------------/
PUNCTUAL RESULTS FILE = (*@\color{PantoneRed}<FileName>@*)
ABSCISSAE OF SPECTRUM PRINTOUT POINTS = (*@\color{PantoneRed}<X1;X2;X3;X4;X5>@*)
ORDINATES OF SPECTRUM PRINTOUT POINTS = (*@\color{PantoneRed}<Y1;Y2;Y3;Y4;Y5>@*)
\end{lstlisting}
\end{CommentBlock}

\item \textbf{Through an external ASCII file:}

  The $X$ and $Y$ can also be defined in an external ASCII file which follows the format
  of DAT outputs of the module {\scshape SMESH} in \salome{}. For a list of nodes, it
  follows this structure:

\lstset{language=TelemacCas,
        basicstyle=\scriptsize\ttfamily}
\begin{lstlisting}[frame=trBL]
<Number of Points> 0
<Point Id> <X< <Y> <Z>
...
\end{lstlisting}

The name of the file is referenced using the \textit{FILE WITH COORDINATES OF SPECTRA
  TO WRITE} keyword.
For example:

\begin{CommentBlock}{Keywords to write a spectra using method 2:}
\lstset{language=TelemacCas,
        basicstyle=\scriptsize\ttfamily}
\begin{lstlisting}[frame=trBL]
/--------------------------------------------------------------------/
/ WRITING SPECTRA
/--------------------------------------------------------------------/
PUNCTUAL RESULTS FILE = (*@\color{PantoneRed}<FileName>@*)
FILE WITH COORDINATES OF SPECTRA TO WRITE =
(*@\color{PantoneRed}<FileName>@*)
\end{lstlisting}
\end{CommentBlock}


\end{enumerate}

\begin{WarningBlock}{Warning:}
  If both methods are used in the same steering file, the simulation will be terminated
  and an error message will be printed.
\end{WarningBlock}

This result file is a SERAFIN formatted file. It first includes all the data about the
spectral mesh geometry, then the names-codes of displayed points. This name-code is of
the type: \textit{Fa\_PT2Db}, where \textit{a} denotes the point's sequence order number
within the list of coordinatesand \textit{b} denotes the number of the closest 2D point
to the specified position. Subsequently, for each graphic printout, it contains the time
and the value of the directional spectrum of wave action for each pair (direction,
frequency) in the spectral mesh.

The keywords \textit{PERIOD FOR GRAPHICS PRINTOUTS }and \textit{NUMBER OF FIRST
  ITERATION FOR GRAPHICS PRINTOUTS}
are shared by the two results files; thus, the printouts are synchronous for either file.

Finally, there is another KEYWORD to store punctual results, it is called \textit{1D
  SPECTRA RESULTS FILE}. The difference with the previous one is that the spectral
energy is summed over directions and that the file is stored in \textit{ASCII} format.
The points where information are stored are defined by the same variables.


\subsection{ The printout listing}

This file contains all the messages as generated by \tomawac during the computation.
It is the main report of a \tomawac run. Its content depends on the value of the
following keyword:

\begin{itemize}
\item  \textit{PERIOD FOR LISTING PRINTOUTS}: this sets the time between two time steps
  of message transmission. This value is given in terms of the number of iterations.
  For example, the following sequence:
\end{itemize}

\textit{TIME INCREMENT} = 30.    

\textit{PERIOD FOR LISTING PRINTOUTS}= 2 

will result in a print in the output listing every 60 seconds of simulation.

\subsection{ The User FORTRAN file}

This User FORTRAN file name is specified in the steering file through the character
keyword: \textit{FORTRAN FILE}. Notice that if the user gives a directory for that
keyword, all the fortran file from this directory will be used.  

The FORTRAN contains all the user-modified \tomawac subroutines as well as the
specifically developed routines for that computation.

This file is compiled and linked during run time in order to generate the executable
being used for the simulation.


\subsection{ The auxiliary files}

 Other input/output files may be used by \tomawac.

\begin{itemize}
\item  A binary data or results file: its name is specified through the character
  keyword \textit{BINARY FILE 1} (Channel unit No. 24).
\item  A formatted data or results file: its name is specified through the character
  keyword \textit{FORMATTED FILE 1} (Channel unit No. 26).
\end{itemize}

These files can be used either for supplying data to the program or for allowing data
to be processed that are not available in the standard results files; obviously, the
user must manage the read and write operations of these files within the FORTRAN program.


\subsection{ The dictionary file}

This dictionary file contains all the information about the keywords (French/English
name, default values, type). This file can be viewed in a text editor by the user,
but it must not be modified in any way.


\subsection{ The libraries}

At the beginning of a computation, the main user-written FORTRAN routine is compiled,
then linked in order to generate the executable program that is subsequently run.

 The following libraries are used during the link editing operation:

\begin{itemize}
\item  \tomawac library: contains the specific subroutines of the \tomawac computation
  model.

\item  telemac libraries: contains the specific subroutines of the TELEMAC-2D or
  TELEMAC-3D computation model.

\item  utils library: contains a number of utility subroutines, such as the partitioning
  and parallel library.

\item  damocles library: contains the subroutines handling the steering file reading.

\item  hermes library: contains the subroutines that permits to read and write data
  according to the file format chosen.

\item  gaia library: contains the subroutines that permits to simulate sediment
  transport

\item  api library : contains the subroutines that permits to the modulus to communicate

\item  BIEF library: contains the computation modules related to the finite
  element-typed operations (operations on both matrixes and vectors). This library is
  shared by all the simulation models as developed by the LNHE within the TELEMAC
  structure (BIEF means "BIbliothèque d'Eléments Finis", i.e. Finite Element Library).
\end{itemize}



\section{ Files standard}

Almost all files that were in Serafin format in previous versions of \tomawac, have
been given a key-word for the file format.

If the name of the file is: ``GEOMETRY FILE'' (``FICHIER DE GEOMETRIE''), the new
keyword will be: ``GEOMETRY FILE FORMAT'' (``FORMAT DU FICHIER DE GEOMETRIE'').

 This format is given in 8 characters. Three choices are possible so far:

\begin{itemize}
\item  `SERAFIN `(do not forget the space at the end): it is the default standard
  within the TELEMAC processing chain. The format is recognized by the FUDAA PRE-PRO
  graphics post-processor. The RUBENS graphics post-processor reads the SERAFIN format
  as well, but it won't be developed anymore and it is bound to disappear. The SERAFIN
  file format is described in detail in \ref{se:desformat}.

\item  `SERAFIND': Serafin format, but with double precision. Can be used for a more
  accurate ``computation continued'' or for more accurate validations. Neither FUDAA
  PRE-PRO nor Rubens can read this format.

\item  `MED  `: this is an EDF-CEA format used in the Salomé platform, that enables to
  use the post-processors of this platform. It is based on hdf5. This new format is not
  activated if you use the default subroutine med.f provided, which is mostly void. If
  you take instead the file med.edf and rename it med.f, med formats will be available,
  but two additional libraries are necessary to use this format and have to be specified
  in the systel.ini file. Full instructions will be given in further releases, this is
  so far for internal use at EDF.
\end{itemize}

A new file structure has been added to library BIEF for simplifying the opening/closing
and reading/writing operations with these file formats, as well as for simplifying the
coupling between programmes,. The description of this file structure and of the
operations on those files are given in \ref{se:filestruct}

As specified in section7.2.7, a fourth binary format exists, which is specific to
\tomawac and is used only for saving the results when they are used to initialize a
next computation. This binary file format cannot be read by the RUBENS post-processor,
or by FUDAA PRE-PRO graphics post-processor.


\section{ Bathymetry data }
\label{se:bathydata}
 The bathymetry information can be supplied to \tomawac at two levels:

\begin{itemize}
\item  Directly in the geometry/mesh file by a bathymetry value being assigned to each
  node in the mesh. In this case the bathymetry data have been processed previously,
  running the STBTEL module or mesh generator. For example, STBTEL reads the information
  from one or more bottom topography files (up to 5 files) and performs an interpolation
  at every point within the domain;

\item  In the form of an irregular pattern of spot heights without any necessary
  relation to the mesh nodes, during the \tomawac computation. The interpolation is
  then performed directly by \tomawac with the same algorithm as used by STBTEL. The
  bathymetry file name is given by the character keyword \textit{BOTTOM TOPOGRAPHY
    FILE. }Unlike STBTEL, \tomawac only handles one bottom topography file. The file
  can be in SINUSX format or can consist of three columns X,Y,Z.
\end{itemize}

\tomawac also provides an opportunity to carry out a smoothing of the bathymetry in
order to get a more consistent geometry. The smoothing algorithm can be iterated several
times in order to achieve more or less extensive smoothing. The number of iterations
is set using the keyword \textit{BOTTOM SMOOTHINGS} and is carried out within the
USER\_TOM\_CORFON subroutine. This keyword's default value is 0. (also refer to the
programming of the CORFON user subroutine in \ref{se:corfon}).

NOTE: the bathymetry data should preferably be supplied to \tomawac in the form of
water depth and not of water height. If necessary, a conversion can be performed in
the USER\_TOM\_CORFON subroutine.


