\chapter{Organisation of the \telemacsystem{} activity}

\section{Structure of the \telemacsystem{} activity}

\subsection{General Organisation}

The \telemacsystem{} is managed by a Consortium composed of two main instances:
\begin{itemize}
\item The Steering Committee,
\item The Technical and Scientific Committee.
\end{itemize}

The main functions of the Steering Committee are as follows:
\begin{itemize}
\item To seek advice from the Technical and Scientific Committee and so to
  decide on the main strategic technical, technological and scientific focuses
  of the Consortium for the medium-to-long term.
\item To seek advice from the Technical and Scientific Committee and so to
  decide, choose and prioritise action items to be included in the Development
  Plan, whether the action items are proposed by the Members of the Consortium
  or by third Parties.
\item To define modes of implementation of the Development Plan and modes of
  integration in the SOFTWARE of the deliverables of the Development Plan
  calling on resources either within the Members of the Consortium or on the
  Experts in Confidence,
\item To make sure that the deliverables of the Development Plan are of desired
  quality, including but not limited to compatible source code, documentation,
  cases for benchmarking and technical approval.
\item To identify, define and validate the strategy for extracting added value
  from the Prior Knowledge and promote the inclusion of Prior Knowledge in the
  software whether it resides within the Consortium or not.
\item To ensure that the present Membership Agreement is performed properly
  and, if necessary, propose endorsements aimed at improving operation of the
  Membership Agreement.
\item To grant/revoke voting rights to Member Representatives.
\item To include/exclude a Member to/from the Consortium subject to the
  duration and termination of a Member's membership.
\end{itemize}

The main functions of the Chairperson of the Steering Committee are as follows:
\begin{itemize}
\item To gather action items, feature wish lists and developments provided by
  Member Representatives and third Parties and to seek development ideas from
  other open source software communities.
\item To convene and preside over the meetings of the Steering Committee.
\item To formalise and report minutes of the meetings of the Steering
  Committee.
\item To formalise and report the Development Plan.
\end{itemize}

The main functions of the Technical and Scientific Committee are as follows:
\begin{itemize}
\item To advise the Steering Committee in its scientific and technical focuses.
\item To diagnose for the Steering Committee the admissibility of the
  developments supplied to the Steering Committee by the Members of the
  Consortium or by third Parties.
\item To enlighten the Steering Committee on the prioritisation to be done and
  the benchmarking and technical approval plans to be implemented.
\end{itemize}

The main functions of the Chairperson of the Technical and Scientific Committee
are as follows:
\begin{itemize}
\item To convene and preside over the meetings of the Representative Members of
  the Technical and Scientific Committee.
\item To formalise and report minutes of the meetings of the Representative
  Members of the Technical and Scientific Committee and gather appropriate
  references, technical notes or scientific evidences supporting its
  conclusions.
\item To synthesise answers to requests submitted via the Chairperson of the
  Steering Committee.
\end{itemize}

\section{Step in the \telemacsystem{} activities}

The \telemacsystem{} activity is divided in three main categories directing the
life of a version of the software, from its creation to its exploitation and
finally to its archiving (See Development Plan in Chapter~\ref{devplan} for
definition of those notions):
\begin{itemize}
\item \textbf{Development activity}: concerns any modification on the code, its
  documentation and the associated tools, whether it is an evolution, a
  correction or a reorganisation.
\item \textbf{Verification and validation activity}: every development must
  pass through the verification and validation tests. A major version or
  production version (see Chapter~\ref{devplan}) must run all the cases.
\item \textbf{Exploitation activity}: concerns the distribution, workshops,
  support, maintenance of the different versions and removal of exploitation,
  as well as the archiving.
\end{itemize}

\subsection{Responsibilities}

The affectation for the different responsibilities are given in the document
``Nominative list of the people of \telemacsystem'' in
Appendix~\ref{peopleoftelma}.

\section{Development activity}
\label{devact}

A development represents any intervention on the source code of the
\telemacsystem, its documentation or its tools (i.e.\ any element under the
configuration, see the development plan in Chapter~\ref{devplan}). The
development on the code can be subdivided into different categories:
\begin{itemize}
\item \textbf{Evolution maintenance}: the evolution maintenance changes the
  software to break some of the limitation it has (performance, strength,
  precision, adaptation to demands\ldots) or to add new functionalities to
  answer new demands or requests.
\item \textbf{Corrective maintenance}: the corrective maintenance removes a bug
  in the code, the examples or the documentation.
\item \textbf{Adaptive maintenance}: the adaptive maintenance adapts the
  software when its environment changes (operating system, compilers, new type
  of clusters\ldots) to ensure that the software still works in that
  environment.
\end{itemize}

Those three categories imply an integration process in the development version
of the \telemacsystem.

The main actors are:
\begin{itemize}
\item The development team from each member of the consortium.
\item The developer from outside firms.
\item The Open-source community, academic and industrial partners.
\end{itemize}
The process attached to those activities are described in the development plan
in Chapter~\ref{devplan}.

\subsection{Processes, tools and rules}

The processes to follow for each development activities are described in the
development plan in Chapter~\ref{devplan}. Following those processes ensure
traceability, non-regression, portability, verification, validation and
re-usability of the functions developed and integrated in \telemacsystem.

Those rules are based on the following principles:
\begin{itemize}
\item Keep the programming coherent with the existing one.
\item Always base yourself on the latest version of the code (the development
  version) when creating new files.
\item Communicate with the developer community on the evolution of the code
  structure.
\end{itemize}

Every new development must have an issue and/or merge request opened on the
\telemacsystem{} GitLab project page
(\url{https://gitlab.pam-retd.fr/otm/telemac-mascaret}), may it be an
evolution, corrective or adaptive maintenance in order to organize and
prioritize the developments done in the \telemacsystem{}.\\

All these developments must be validated by the associated CH and the PM\@.

\subsection{Evolution maintenance}

The developments realised for any maintenance are under the responsibility of
the developer who was given this assignment by the CHs. They are handled by
following the processes described in the development plan in
Chapter~\ref{devplan}.\\

The final integration of the development of a maintenance is in charge of the
associated CH and must follow the process described in the development plan in
Chapter~\ref{devplan}.

\section{Verification and validation activity}
\label{vv}
The verification and validation activity aims to verify the implementation on
an algorithmic level or on the complete model using verification and validation
test cases (even if a small part of the code was impacted as it can be for the
corrective and adaptive maintenance). It is complementary with the test
realised during the implementation (See the development plan in
Chapter~\ref{devplan}). We distinguish them as follow:
\begin{itemize}
\item \textbf{Verification}: The verification aims to establish that the code
  solves efficiently the mathematical problem it is supposed to resolve (and
  that is described in the specification of the concern algorithm). The
  verification focuses on the quality of the implementation from a computer
  science and theoretical point of view. The answer to the verification process
  is a binary one: the code does or does not respond to the demands. To do that
  verification, an analytical solution of the model or a solution based on
  measurements with the possible addition of source terms to the equation is
  used. The process is then most of the time to compare the numerical solution
  with a reference solution on meshes with increasing mesh refinement. This
  verifies that the algorithm converges in mesh and gives the level of
  precision associated.
\item \textbf{Validation}: The validation aims to see to what level the code is
  able to answer to a physical problem in a given application domain.  The
  validation focuses on the ability of the code to predict event from a
  physical point of view. The answer to this process is not binary. The
  evaluation is made through a comparison between a numerical solution and an
  experimental solution. We distinguish the validation with separated effects,
  which focuses on an isolated physical phenomenon and the full validation
  which focuses on an industrial problem (which can be simplified to get access
  to experimental measurements) composed of multiple physical phenomena.
\item \textbf{Qualification}: The qualification aims to establish the perimeter
  of action of the code by using heavily validated complexed cases. This action
  is not done by the developers, it is in charge of main users. The creation of
  a methodology note is part of the qualification process. The qualification
  can keep track of the performance of the code as well as the non-regression
  of the test cases.
\end{itemize}

\section{Release and distribution activity}

\subsection{Release, reception, removal}

Between the different versions of the \telemacsystem, we distinguish the
release of a major version or production version from the other versions
(minor, development release).\\

The different types of version are given in the development plan in
Chapter~\ref{devplan}.\\

The release of a new major version of the \telemacsystem{} is decided by the
chief of department hosting the project.\\

The development version of the \telemacsystem{} is not formally released.
Nevertheless it is compiled and test cases are run periodically. The latest
development version is available on the official GitLab repository
(\url{https://gitlab.pam-retd.fr/otm/telemac-mascaret}).\\

The process for the removal of older stable version is described in the
development plan in Chapter~\ref{devplan}.

\subsection{Distribution}

\textbf{Distribution in EDF R\&D}

The major and minor versions of the \telemacsystem{} are installed on the MNEMO
data storage solution on an area dedicated to the project. The DSIT handles
the backup of these data.\\

The test cases are available (at least in read-only mode) on MNEMO. The
verification and validation is done on HPC clusters based in EDF R\&D (France).
The environments on which the code is verified and validated are listed in the
release note.\\

\textbf{Distribution in EDF}

The distribution outside of the R\&D is under the responsibility of the
\telemacsystem{} Chain Handler. The environment guide written by the Chain
Handler contains a help for the installation of the code.\\

\textbf{Open source distribution}

All the stable versions of the \telemacsystem, the development version and the
tools are under open source license (GPL for the \telemacsystem{} modules
except for the \bief{} which is under LGPL) and are available on the website
\url{http://www.opentelemac.org} which is the main access for all information
about the \telemacsystem.\\

\textbf{Protection}

The software is protected in France by the label \tel{} since 05/02/96 with the
number INPI 96/623 748.

\subsection{Relation with the \telemacsystem{} users}

The services handled by the \telemacsystem{} project and available to users
are:
\begin{itemize}
\item A technical assistance: forum, bug tracker, guides, documentation.
\item Meeting with users: internally in EDF or externally (User Conference)
\end{itemize}

All those services are available on the \telemacsystem{} website
(\url{http://www.opentelemac.org}).
The developers responding to the forum are not bound to respond to the forum,
they do that on their available time.\\

\textbf{Technical assistance}

The organisation of the technical assistance is divided into two levels:
\begin{itemize}
\item \textbf{First level assistance}\\ This assistance can concern either the
  installation or the usage of the software. This assistance can lead to a
  demand for an evolution of the code or the discovery of a bug. The
  formalisation of those demands allows:
  \begin{itemize}
  \item A better repartition of resources by the project.
  \item The follow-up by the user of the demand and allowing him to interact
    with the developer handling the demand.
  \end{itemize}
\item \textbf{Second level assistance}\\ This assistance is done by the
  development team of the \telemacsystem{}. This is an anticipated action to
  create documentation and didactic software (user and developer documentation,
  online documentation, tutorials, FAQs\ldots).
\end{itemize}

The assistance activity can lead to a development if the problem is deemed
worth it.\\

The actors of this activity are:
\begin{itemize}
\item Users: they are telling us their needs, helping in the specification and
  the validation of the code. They can also respond to some questions on the
  forum.
\item People allowed to submit development demands, more specifically the
  development team (see the development plan in Chapter~\ref{devplan}): they
  answer to the forum and GitLab issues\@. They can subcontract some of the
  work.
\item The Kernel Handler: handles the GitLab repository, its maintenance and
  its good use.
\end{itemize}

\textbf{Meeting with users}

The meeting with user orbits around the following element:
\begin{itemize}
  \item \textbf{the \telemacsystem{} User Conference}
\end{itemize}

\subsection{Relation with partners}

The partners (like for example the members of the \telemacsystem{} consortium)
must follow the same rules for the development as described in the development
plan in Chapter~\ref{devplan}.

\section{Control of sub-contract}

Any developer from the development team can sub-contract some of his
development like:
\begin{itemize}
\item The corrective, adaptive, evolution maintenance.
\item Action on the documentation.
\item Action of verification and validation.
\end{itemize}

In this case the process and the rules to follow are the same as an internal
development described in the development plan in Chapter~\ref{devplan}. The
developer ordering this sub-contract is in charge of the follow-up of the work
and its future maintenance.

Other actions can be sub-contracted:
\begin{itemize}
\item Action on the website.
\item Action of support.
\end{itemize}
