\chapter{The inputs / outputs}

\section{Preliminary comments}

In a computation, the \telemac{3D} code uses a number of input and output files,
some of which are optional. Most of these files are similar or identical to
their counterparts in \telemac{2D}.

The input files are:

\begin{itemize}
\item The steering file (mandatory), which contains the "configuration" of the
simulation,

\item The geometry file (mandatory), which contains the information regarding
the mesh,

\item The boundary conditions file (mandatory), which contains the description
of the type of each boundary,

\item The FORTRAN file, which contains the specific subroutines of the
simulation (modified TELEMAC subroutine or specifically created),

\item The bottom topography file, containing the elevation of the bottom.
Generally, the topographic data are already available in the
geometry file and the bottom topography file is generally not used,

\item The liquid boundaries file, which contains the information on the
imposed values at liquid boundaries,

\item The previous computation file, which provides the initial state of the
calculation in the case of a restart calculation,

\item The reference file, which contains the calculation of "reference" used
for the validation process,

\item The stage-discharge curves file, which contains the information on the
imposed values at liquid boundaries in case of height / flow rate law,

\item The sources file, which contains the information regarding the sources,

\item The binary boundary data file, which contains boundary conditions data
varying in time and space,

\item The time series coordinates file which contains coordinates and periods of
time where time series are extracted and written in the 2D and 3D time series
files,

\item The sections input file, which contains the description of the control
sections of the model (sections across which the flowrate is computed).
\end{itemize}

The output files are:

\begin{itemize}
\item The 3D result file, which contains the graphical results associated to
the 3D mesh,

\item The 2D result file, which contains the graphical results associated to
the 2D mesh,

\item The output listing, which is simulation report. In case of difficulty in
performing a calculation, the user can request the printing of additional
information by activating the logical keyword \telkey{DEBUGGER} (default = 0).
\telkey{DEBUGGER = 1} provides the calling sequences of subroutines from the
main program \telfile{TELEMAC3D}. This technique is useful in case of
critical computation crash to identify the responsible subroutine,

%\item The file for Scope, which is an additional test file available for the
%user,
%
\item The restart file, allowing to perform a restart computation without
information loss,

\item The 2D and 3D time series files which contain time series at locations
defined in the time series coordinates file,

\item The sections output file, which contains the results of the
``control sections'' computation.
\end{itemize}

In addition, the user may have to manage additional files:

\begin{itemize}
\item The binary data files 1 and 2, in input,

\item The formatted data files 1 and 2, in input,

\item The binary results file, in output,

\item The formatted results file, in output.
\end{itemize}


\subsection{Binary files format}
\label{sec:binfile}
The binary files used within the TELEMAC system can have various formats. The
most common format is the SERAFIN format (also known as SELAFIN, for
misidentified historical reasons) which is the standard internal TELEMAC system
format (described in Appendix \ref{sec:srffmt}). This SERAFIN format can be
configured so as to store the real-typed values in single or double precision.
The other available format is the MED format which is used by the SALOME
platform jointly developed by EDF and the French atomic energy commission
(CEA).
The full description of the MED format is available on the SALOME website
http://www.salome-platform.org.
%In the current TELEMAC release, this format is restricted to EDF
%internal use.

Depending on the specified format, the binary files may be treated by different
software. However, in the current release of TELEMAC, only the SERAFIN single
precision format can be read by the post-processing tools like RUBENS
or Blue Kenue.
It is not the case for Paraview or FUDAA-PREPRO.
%Tecplot?

The selection of the format of a file is done by the corresponding keyword.
Thereby, the keyword \telkey{GEOMETRY FILE FORMAT} specifies the format of the
geometry file. Those keywords may take 3 different values (as an 8 character
string): the value 'SERAFIN ' corresponds to the standard single precision
SERAFIN format, which is the default and recommended value (do not forget the
space in last position). The 'SERAFIND' value corresponds to the double
precision SERAFIN format which allows to increase the precision of the results,
especially in case of a restart or reference file. Finally, the 'MED' value
corresponds to the HDF5 MED format in double precision.

The keywords involved are:

\begin{itemize}
\item \telkey{2D RESULT FILE FORMAT},

\item \telkey{2D TIME SERIES FILE FORMAT},

\item \telkey{3D RESULT FILE FORMAT},

\item \telkey{3D TIME SERIES FILE FORMAT},

\item \telkey{BINARY ATMOSPHERIC DATA FILE FORMAT},

\item \telkey{BINARY DATA FILE 1 FORMAT},

\item \telkey{BINARY RESULTS FILE FORMAT},

\item \telkey{FILE FOR 2D CONTINUATION FORMAT},

\item \telkey{GEOMETRY FILE FORMAT},

\item \telkey{PREVIOUS COMPUTATION FILE FORMAT},

\item \telkey{REFERENCE FILE FORMAT},

\item \telkey{RESTART FILE FORMAT},

\item \telkey{TIDAL MODEL FILE FORMAT}.
\end{itemize}

\section{The steering file}

The steering file contains all the data about the selection of computational
options (physical, numerical, etc.). It is an ASCII file which can be generated
either through the FUDAA-PREPRO or EFICAS software or directly using a text
editor. In a way, it serves as the computation dashboard. It includes a set of
keywords to which values are assigned. If a keyword does not occur in that
file, then \telemac{3D} will assign it the default value as defined in the
dictionary file. If such a default value is not defined in the dictionary, then
the computation will be interrupted with an error message. For instance, the
command \telkey{TIME STEP = 10.0} specifies that the computation time step
value is 10 seconds.

\telemac{3D} reads the steering file at the beginning of the computation.

Both dictionary file and steering file are read using the \telfile{damocles}
library which is included in the \tel chain. It is therefore necessary,
when generating the steering file, to observe the \telfile{damocles} syntax
rules (what is performed automatically if the file is generated using
FUDAA-PREPRO).

The writing rules are as follows:

\begin{itemize}
\item The keywords can be of the Integer, Real, Logical or Character type,

\item The sequence order in the steering file is of no importance,

\item Several keywords can be on the same line,

\item Each line cannot exceed 72 characters. However, one can start a new line
as many times as one wishes provided that the keyword name is not astride two
lines,

\item For the array-types keywords, the character separating successive values
is the semicolon. For example:

\begin{lstlisting}[language=TelemacCas]
PRESCRIBED FLOWRATES =   10.0;20.0
\end{lstlisting}

\item The symbols ":" and "=" can both be used in order to separate a keyword
name and its value. They can be either preceded or followed by any number of
blanks. The value itself can occur on the following line. For example:

\begin{lstlisting}[language=TelemacCas]
TIME STEP =   10.
\end{lstlisting}

or

\begin{lstlisting}[language=TelemacCas]
TIME STEP : 10.
\end{lstlisting}

or else

\begin{lstlisting}[language=TelemacCas]
TIME STEP =
10
\end{lstlisting}

\item The characters occurring between two "/" on one line are taken as
comments. Likewise, those characters occurring between a "/" and a line ending
are considered as comments. For example:

\begin{lstlisting}[language=TelemacCas]
/ k-epsilon model
VERTICAL TURBULENCE MODEL = 3
\end{lstlisting}

\item A line beginning with a "/" in a first column is wholly considered as a
comment, even though there is another "/" on the line. For example:

\begin{lstlisting}[language=TelemacCas]
/ The geometry file is ./mesh/geo
\end{lstlisting}

\item Writing the integers: do not exceed the maximum size allowed by the
machine (in a machine with a 32 bit architecture, the extreme values range from
-2~147~483~647 to +2~147~483~648. Do not insert any blank between the sign
(optional for the sign +) and the number. A point after the end of the number
is tolerated,

\item Writing the reals: the point and the comma are accepted as a decimal
separator, as well as the FORTRAN formats E and D. ( 1.E-3  0.001  0,001  1.D-3
represent the same value),

\item Writing the logical values: the values 1, OUI,  YES,  .TRUE.,  TRUE,  or
VRAI on the one hand, and 0, NON,  NO,  .FALSE.,  FALSE, or FAUX on the other
hand, are accepted,

\item Writing the character strings: the strings including blanks or reserved
symbols ("/",":", "=", "\&") should be inserted between quotes ('). The value
of a character keyword may contain up to 144 characters. As in FORTRAN, the
quotes included in a string should be doubled. A string may not begin or end
with a blank. For example:

\begin{lstlisting}[language=TelemacCas]
TITLE = 'COASTAL ENVIRONMENT STUDY'
\end{lstlisting}

\end{itemize}

In addition to the keywords, a number of pseudo instructions or metacommands
which are interpreted during the sequential reading of the steering file can
also be used:

\begin{itemize}
\item The \telkey{\&FIN} command specifies the file end (even though the
file is not over). Thus, some keywords can be disabled simply by placing them
behind that command so that they can easily be reactivated later on. However,
the computation keeps running,

\item The \telkey{\&ETA} command prints the list of keywords and the
values which are assigned to them when \telfile{damocles} meets that command.
That display will take place at the head of the output listing,

\item The \telkey{\&LIS} command prints the list of keywords. That
display will take place at the head of the output listing,

\item The \telkey{\&IND} command prints the detailed list of keywords.
That display will take place at the head of the output listing,

\item The \telkey{\&STO} command causes the program to be halted, whereas
the computation does not keep running.
\end{itemize}

\section{The geometry file}

It is the same file as that used by \telemac{2D}.
This is a binary file.
The format of this file is given by the keyword \telkey{GEOMETRY FILE FORMAT}.
There are 3 possibilities:
\begin{itemize}
\item SERAFIN: classical single precision format in \tel,
\item SERAFIND: classical double precision format in \tel,
\item MED: MED double precision format based on HDF5.
\end{itemize}

If it is a SERAFIN-formatted binary file, it can be read by
Paraview, FUDAA-PREPRO, Tecplot 360, Blue Kenue or RUBENS
and is generated either by SALOME-HYDRO, Janet, Blue Kenue or MATISSE,
or by the \stbtel module (from the file(s) originating from of mesh generator).
The SERAFIN format structure is described in Appendix \ref{sec:srffmt}.

If it is a MED-formatted binary file, it can be read by
Paraview only and can be generated by SALOME-HYDRO
or by the \stbtel module (from the file(s) originating from of mesh generator).

That file contains all the data about the two-dimensional mesh (see in
subsection \ref{sec:2dmesh}).
It includes the number of points in the mesh (\telfile{NPOIN2}
variable), the number of elements (\telfile{NELEM2} variable), the number of
vertices per element (\telfile{NDP} variable), the \telfile{X} and \telfile{Y}
arrays containing the co-ordinates of all the points and, lastly,
the \telfile{IKLE} array containing the connectivity table.

That file may also contain bathymetric and bottom friction data for each point
in the mesh.

NOTE: \telemac{3D} retrieves the geometry data at the beginning of the 2D result
file. Any computational 2D result file can therefore be used as a geometry file
when a further simulation on the same mesh is desirable.

That file name is provided using the keyword: \telkey{GEOMETRY FILE} and its
format is specified by the keyword \telkey{GEOMETRY FILE FORMAT} (default
value: 'SERAFIN '), see subsection \ref{sec:binfile}.


\section{The boundary conditions file}

It is the same file as that used by \telemac{2D}. It is a formatted file which
is generated automatically by most of the mesh generators compatible with \tel
or \stbtel and can be amended using FUDAA-PREPRO or a text editor.
Each line in that file is dedicated to a point at the 2D mesh boundary.
The edge point numbering is that of the file lines;
it first describes the domain outline in the counter clockwise direction
from the bottom left-hand point (point where the sum of horizontal co-ordinates
of which is minimum), then the islands in the clockwise direction.

For a thorough description of that file, refer to the specific subsection
\ref{sec:bndfile}.

That name of file is provided using the keyword: \telkey{BOUNDARY CONDITIONS
FILE}.


\section{The FORTRAN file}

The FORTRAN file (a single file or a directory containing one or several file(s))
may include a number of subroutines (so-called "user"
subroutines) available under the \telemac{3D} tree structure which the user can
modify as well as those subroutines which have been specifically developed for
the computation.

The user subroutines from the various libraries used by \telemac{3D} are listed
in Appendix \ref{sec:usrsub}. Since \telemac{3D} is available in Open Source, every
subroutine can be freely used. Every user subroutine copied into the user
FORTRAN file is automatically substituted for the same named subroutine
occurring in the \telemac{3D} compiled libraries.

Upon the creation and every amendment of the FORTRAN file, a new executable
program is generated (compilation and link) for the simulation.

That file name or directory name is provided using the keyword
\telkey{FORTRAN FILE}.


\section{The liquid boundaries file}

It is an ASCII file enabling the user to specify time-varying boundary
conditions values (flow rate, depth, velocity, tracer concentration) at all the
liquid boundaries. That file can be generated under the FUDAA-PREPRO software
interface.

For a thorough description of that file, refer to the specific subsection
\ref{sec:liqbnd}.

That name of file is provided using the keyword: \telkey{LIQUID BOUNDARIES
FILE}.


\section{The previous computation file}
\label{sec:previousfile}

It is a \telemac{3D} result file which is used for initializing a new
computation. In order to activate the optional use of that file, the keyword
\telkey{COMPUTATION CONTINUED} should be activated (default = NO).
In order to specify the
previous computation file, its name should be stated through the keyword:
\telkey{PREVIOUS COMPUTATION FILE}. The initial conditions of the new
computation are defined by one recorded time step in the previous
computation file given by the keyword \telkey{RECORD NUMBER FOR RESTART}.
By default, its value is -1 which means it is the last time step
to start from in the \telkey{PREVIOUS COMPUTATION FILE}.
The whole set of data from the steering file is read and it
makes possible to redefine or amend the variables (time step, turbulence model,
addition or deletion of a tracer\dots).

A computation can also be initialized from a \telemac{2D} result. In order to
activate that option, the \telkey{2D CONTINUATION} keyword
should be validated (default = NO).
The \telemac{2D} result file should then be associated with
the \telkey{FILE FOR 2D CONTINUATION FORMAT} keyword (default
value: 'SERAFIN '), see subsection \ref{sec:binfile}.


\section{The reference file}

During the validation step of a calculation, this file contains the reference
result.
At the end of the calculation, the result of the simulation is compared to the
last time step stored in this file. The result of the comparison is given in
the control printout in the form of a maximum difference in elevation and the
three components of velocity.

The name of this file is given by the keyword \telkey{REFERENCE FILE} and its
format is specified by the keyword \telkey{REFERENCE FILE FORMAT} (default
value: 'SERAFIN '), see subsection \ref{sec:binfile}.


\section{The stage-discharge curves file}

This text file enables the user to configure the evolution of the prescribed
value on specific open boundaries. This file is used when the prescribed
elevation is determined by a discharge-elevation law. The descriptions of the
appropriate laws are given through this file.

See subsection \ref{sec:discharge} for a complete description of this file.

The name of this file is specified with the keyword: \telkey{STAGE-DISCHARGE
CURVES FILE}.


\section{The sections input file}

This ASCII file enables the user to configure the control sections used
during the simulation.

See section \ref{sec:ctrlsc} for a complete description of this file.

The file name is specified with the keyword: \telkey{SECTIONS INPUT FILE}.


\section{The sources file}

This text file enables the user to specify values for time-dependent conditions
for sources (discharge, tracers concentration).

See section \ref{sec:srcfile} for a complete description of this file.

The file name is specified with the keyword: \telkey{SOURCES FILE}.


\section{The binary boundary data file}

This binary file enables the user to specify boundary conditions data (e.g.
elevations, horizontal components of velocity, tracer values) which
may vary in time and space.

See the Python script \telfile{converter.py} in the scripts/python3 directory
with the positional argument generate\_bnd.
It may help to generate such a file from a global model which will be
interpolated on a geometry file for every liquid node given by the boundary
conditions file.

The name of this file is given by the keyword \telkey{BINARY BOUNDARY DATA FILE}
and its format is specified by the keyword
\telkey{BINARY BOUNDARY DATA FILE FORMAT} (default value: 'SERAFIN '),
see subsection \ref{sec:binfile}.

\begin{WarningBlock}{Caution:}
To use the \telkey{BINARY BOUNDARY DATA FILE}, some keywords have to be added
in the steering file:
\begin{itemize}
\item \telkey{PRESCRIBED ELEVATIONS} with at least as many values as the
number of liquid boundaries if the \telkey{BINARY BOUNDARY DATA FILE} contains
elevations at the boundaries,
\item \telkey{PRESCRIBED VELOCITIES} with as at least many values as the
number of liquid boundaries if the \telkey{BINARY BOUNDARY DATA FILE} contains
horizontal components of velocity at the boundaries.
Moreover, the two velocity components $U$ and $V$ must both exist in the file,
\item \telkey{PRESCRIBED TRACERS VALUES} with at least as many values as the
number of liquid boundaries times the number of tracer(s) if the
\telkey{BINARY BOUNDARY DATA FILE} contains tracer(s) values at the boundaries,
\end{itemize}
\end{WarningBlock}


\section{The 3D result file}
\label{sec:3dres}
It is the file into which \telemac{3D} stores the information during the
computation. It is a SERAFIN-formatted file (refer to Appendix
\ref{sec:srffmt}) or a MED-formatted file. It first contains all the data about
mesh geometry, then the names of the stored variables. It also contains, for
each graphic printout and for each mesh point, the values of the various
recorded variables.

Its content varies according to the values of the following keywords:

\telkey{NUMBER OF FIRST TIME STEP FOR GRAPHIC PRINTOUTS}: provided to set from
which time step onwards the information will be stored, in order to prevent too
large files, especially when the computation begins with an uninteresting
transient stage related to the definition of unrealistic initial conditions
(e.g. invariably zero currents). The default value is 0 (writing of the graphic
printouts at the beginning of the simulation).

\telkey{GRAPHIC PRINTOUT PERIOD}: sets the period (in number of time steps) of
printouts in order to prevent a too large file.  Default value is 1 (writing at
every time step). For instance:

\begin{lstlisting}[language=TelemacCas]
TIME STEP  =  60.0
GRAPHIC PRINTOUT PERIOD =  30
\end{lstlisting}

The results will be backed up every ${1,800}^{\rm{th}}$ second, i.e.
${30}^{\rm{th}}$ minute.

\telkey{VARIABLES FOR 3D GRAPHIC PRINTOUTS}: keyword specifying the list of
variables which will be stored in the result file. Each variable is identified
by means of a name from the list below.

\begin{itemize}
\item Z elevation (m),

\item U velocity along $x$ axis (m/s),

\item V velocity along $y$ axis (m/s),

\item W velocity along $z$ axis (m/s),

\item TA concentrations for tracers (TA1 for the ${1}^{\rm{st}}$ one, TA2 for the
${2}^{\rm{nd}}$ one\dots ),

\item NUX viscosity for U and V along $x$ axis (m${}^{2}$/s),

\item NUY viscosity for U and V along $y$ axis (m${}^{2}$/s),

\item NUZ viscosity for U and V along $z$ axis (m${}^{2}$/s),

\item NAX viscosity for tracers along $x$ axis (m${}^{2}$/s),

\item NAY viscosity for tracers along $y$ axis (m${}^{2}$/s),

\item NAZ viscosity for tracers along $z$ axis (m${}^{2}$/s),

\item RI Richardson number in case of a mixing length model,

\item K turbulent energy for $k$-$\epsilon$ model (J/kg),

\item EPS dissipation of turbulent energy (W/kg),

\item DP dynamic pressure (multiplied by DT/RHO),

\item PH hydrostatic pressure (Pa),

\item RHO relative density,

\item P1 private variable 1,

\item P2 private variable 2,

\item P3 private variable 3,

\item P4 private variable 4,

\item US : Stokes velocity along $x$ axis (m/s),

\item VS : Stokes velocity along $y$ axis (m/s),

\item WS : Stokes velocity along $z$ axis (m/s),

\item CSO2 oxygen saturation concentration (mgO2/l) if coupling with \waqtel
and using O2 or EUTRO model,

\item O2SAT percentage of oxygen saturation (ratio oxygen concentration over
oxygen saturation concentration $C_s$ in \%) if coupling with \waqtel and using
O2 or EUTRO model.
\end{itemize}

That file name is provided using the keyword: \telkey{3D RESULT FILE} and its
format using \telkey{3D RESULT FILE FORMAT} (default value: 'SERAFIN '). Stored
variables by default are `Z, U, V, W'.

Using the private arrays requires updating the keyword \telkey{NUMBER OF
PRIVATE ARRAYS} in order to perform the required memory allocations (see
section \ref{sec:privarray}).

\section{The 2D result file}

It is the file into which \telemac{3D} stores the specifically two-dimensional
data during the computation (such as the free surface, the depth-averaged
horizontal components of velocity and the depth-averaged tracers). It has a
SERAFIN format. The free surface and the horizontal components of velocity will
then physically correspond to the same data as those being supplied by
\telemac{2D}. The obtained values, however, may be different from an analogue
computation being directly made with \telemac{2D} when the flow is specifically
three-dimensional.

Its content varies according to the values of the following keywords:

\telkey{NUMBER OF FIRST TIME STEP FOR GRAPHIC PRINTOUTS}: same keyword as that
described in section \ref{sec:3dres}.

\telkey{GRAPHIC PRINTOUT PERIOD}: same keyword as that described in section
\ref{sec:3dres}.

\telkey{VARIABLES FOR 2D GRAPHIC PRINTOUTS}: keyword specifying the list of
variables which will be stored in the result file. Each variable is identified
by means of a name from the list below.

\begin{itemize}
\item U average velocity along $x$ axis (m/s),

\item V average velocity along $y$ axis (m/s),

\item C celerity (m/s),

\item H water depth (m),

\item S free surface elevation (m),

\item B bottom elevation (m),

\item TA averaged concentrations for tracers (TA1 for the ${1}^{\rm{st}}$ one, TA2 for the
${2}^{\rm{nd}}$ one\dots ),

\item F Froude number,

\item Q scalar discharge (m${}^{2}$/s),

\item I discharge along $x$ (m${}^{2}$/s),

\item J discharge along $y$ (m${}^{2}$/s),

\item M norm of velocity (m/s),

\item X wind along $x$ axis (m/s),

\item Y wind along $y$ axis (m/s),

\item P air pressure (Pa),

\item W friction coefficient,

\item RB non erodible bottom elevation (m),

\item HD thickness of the sediment bed layer (m),

\item EF erosion rate (kg/m${}^{2}$/s),

\item DF deposition flux (kg/m${}^{2}$/s),

\item DZF bed evolution,

\item PRIVE1 private array PRIVE 1,

\item PRIVE2 private array PRIVE 2,

\item PRIVE3 private array PRIVE 3,

\item PRIVE4 private array PRIVE 4,

\item QS  solid discharge (m${}^{2}$/s),

\item QSX  solid discharge along $x$ (m${}^{2}$/s),

\item QSY  solid discharge along $y$ (m${}^{2}$/s),

\item US  friction velocity (m/s),

\item MAXZ  maximum value of the water elevation during the computation (m),

\item TMXZ  time corresponding to this maximum elevation (s),

% 2D variables K, E, D, A, G, L not relevant or not programmed
% see subroutine lecdon_telemac3d with the array SORG2D

%\item K turbulent energy (J/kg),

%\item E dissipation (W/kg),

%\item D turbulent viscosity (m$^2$/s),

%\item A drift along $x$ (m),

%\item G drift along $y$ (m),

%\item L courant number,

\item TAIR  air temperature ($^{\circ}$C),

\item USURF surface velocity along $x$ axis (m/s),

\item VSURF surface velocity along $y$ axis (m/s),

\item WSURF surface velocity along $z$ axis (m/s),

\item MSURF magnitude of surface velocity (m/s),

\item TASURFi concentrations for tracers at the surface,
$i$ is the tracer number,

\item TASURF* concentrations for tracers at the surface from 1 to 9,

\item TASURF** concentrations for tracers at the surface from 10 to 99,

\item CSO2 average oxygen saturation concentration (mgO2/l) if coupling with
\waqtel and using O2 or EUTRO model,

\item O2SAT average percentage of oxygen saturation (ratio oxygen concentration
over oxygen saturation concentration $C_s$ in \%) if coupling with \waqtel and
using O2 or EUTRO model.
\end{itemize}

That file name is provided by means of the keyword: \telkey{2D RESULT FILE} and
its format using \telkey{2D RESULT FILE FORMAT} (default value: `SERAFIN ').
Stored variables by default are `U, V, H, B'.

Using the private arrays requires updating the keyword \telkey{NUMBER OF 2D
PRIVATE ARRAYS} in order to perform the required memory allocations (see
section \ref{sec:privarray}).

If the user is only interested in the values of velocities at the surface and/or
tracer values at the surface rather than the whole 3D domain,
he/she can save them in the \telkey{2D RESULT FILE} with the dedicated variables
with SURF suffix.
That enables to save memory in the \telkey{3D RESULT FILE}.

\section{The time series files}
Time series files are files (in SERAFIN or MED format), in which time series are
written for certain points in the domain of the simulation.
This feature has been introduced since release 8.5.
In this way, one can obtain results with a high time frequency, without using
a large amount of disk space.
The time series file module interpolates data (using linear interpolation)
to the required output coordinates.
\begin{itemize}
\item \telkey{2D TIME SERIES FILE} The name of the output file to write
2D results.
The time and coordinates correspond to those given in
\telkey{TIME SERIES COORDINATES FILE},
\item \telkey{3D TIME SERIES FILE} The name of the output file to write
3D results for time and
coordinates given in \telkey{TIME SERIES COORDINATES FILE},
\item \telkey{2D TIME SERIES FILE FORMAT} The file format of
\telkey{2D TIME SERIES FILE} (single or double precision SERAFIN, or MED,
default = SERAFIN),
\item \telkey{3D TIME SERIES FILE FORMAT} The file format of
\telkey{3D TIME SERIES FILE} (single or double precision SERAFIN, or MED,
default = SERAFIN),
\item \telkey{TIME SERIES COORDINATES FILE} The file which contains the output
times and output coordinates for 2D and 3D \telkey{TIME SERIES FILE}.
The format is described next.
\end{itemize}

The \telkey{TIME SERIES COORDINATES FILE} has following format:
\begin {itemize}
\item On the first line, there are two numbers indicating:
\begin{itemize}
\item The number of time periods for which data is written to the
\telkey{TIME SERIES FILE},
\item The number of coordinates for which output is generated.
\end {itemize}
\item On the next lines, for each output period, there are three numbers
indicating:
\begin{itemize}
\item Start of the output period (in seconds since the start of the computation),
\item End of the output period (in seconds since the start of the computation),
\item Output time interval (in seconds).
\end {itemize}
\item For each coordinate, three numbers and a string, containing respectively:
\begin{itemize}
\item $x$-coordinate of the output point,
\item $y$-coordinate of the output point,
\item Unique ID of each station (in order to easily encounter the output
data in the output files).
This ID will be written to the IKLE array in the output file, which has a size
[NPOINTS $\times$ 1].
Note that the time series file consists of a list of points.
There is no mesh information in a time series file.
Therefore, the length of the IKLE array is the same as the number of points in
the time series file,
\item The name of the station (for better readability of the coordinate file).
\end{itemize}
\item Anything after the list of coordinates is considered comment and ignored
by \tel.
\end {itemize}

An example of a \telkey{TIME SERIES COORDINATES FILE} is given below.

\begin{verbatim}
2 3
0 100 10
500 1000 20
250  175     01 output_point_01
500  175     02 output_point_01
750  175     03 output_point_02

Description of the file
1st line: number of periods and number of points
2nd line: period 1: start time, end time and interval (in seconds)
3rd line: period 2: start time, end time and interval (in seconds)
4th-6th line: output points; x and y coordinates, unique ID, and station name
\end{verbatim}

\section{Results in longitude-latitude}
If the geometry file is given in longitude-latitude, the result files
can also be written with the coordinates in longitude-latitude
by setting the keyword \telkey{RESULT FILE IN LONGITUDE-LATITUDE}
at YES (default value).

\section{The output listing}

It is a formatted file which can be created by \telemac{3D} during the
computation (program launched with the -s option). It contains the report of a
\telemac{3D} running.
The default value of the keyword \telkey{LISTING PRINTOUT} is YES,
which means the listing is written.
If NO is selected, the listing only includes the heading and the
phrase "NORMAL END OF PROGRAM".
It is not recommended to change the default value.
In addition, the options \telkey{MASS-BALANCE} and \telkey{VALIDATION} are inhibited.

Its contents vary according to the values of the following keywords:

\begin{itemize}

\item \telkey{NUMBER OF FIRST TIME STEP FOR LISTING PRINTOUTS}: provided to set at
which time step it is desired to begin printing the data, in order to prevent
too large files.
The default value is 0 (writing of the listing printouts at the
beginning of the simulation),

\item \telkey{LISTING PRINTOUT PERIOD}: sets the period between two time step
printings. The value is given as a time step number.
By default, it is equal to 1, i.e. every time step.
For instance, the following sequence:

\begin{lstlisting}[language=TelemacCas]
TIME STEP = 30.0
LISTING PRINTOUT PERIOD = 2
\end{lstlisting}

prints the output listing every minute of simulation,

\item \telkey{MASS-BALANCE}: if it is requested, the user will get
information about the mass flow (or rather the volumes) and the errors
(primarily linked to the precision achieved by the solvers) of that computation
in the domain. This is not done by default,

\item \telkey{INFORMATION ABOUT MASS-BALANCE FOR EACH LISTING PRINTOUT}: if they are
requested (that is done by default), the user will get, at each time step,
information about the flows within the domain.

The file name is directly managed by the \telemac{3D} launching procedure.
Generally, the file has a name which is created from the name in the steering
file and the number of the processors used in the computation, followed by the
suffix "\telfile{.sortie}".

\end{itemize}

\section{The sections output file}

Rather than reading the flowrates computed on control section(s) in the listing,
the user can store these results in a file with the keyword
\telkey{SECTIONS OUTPUT FILE}.
It is an ASCII file, written by the master in case of parallel run.

%\section{The file for scope}
%
%\telemac{3D} the user an opportunity to retrieve computation variables
%along a 1D profile during a computation. These profiles will then be stored in
%a dedicated file the name of which is defined by the keyword \telkey{FILE FOR
%SCOPE}. For that purpose, the user should program the \telkey{SCOPE} subroutine.
%
%That subroutine is provided to define profiles of computed variables, or other
%variables, as created by a user along a segment with (X1, Y1, Z1) and (X2, Y2,
%Z2) co-ordinates. The user also decides the number of points as distributed
%along that segment. The data is automatically saved as per the SCOPE format at
%all the time steps (for a description of this format, see Appendix of RUBENS
%Version 4.0. Manuel Utilisateur, O. Quiquempoix, EDF report HE-45/95/031/B).
%
\section{The auxiliary files}

Other files can be used by \telemac{3D}:

\begin{itemize}
\item One or two binary data files, as specified by the keywords
\telkey{BINARY DATA FILE 1} and \telkey{BINARY DATA FILE 2}. These files can be
used to provide data to the program, the user has to handle their reading
within the \telkey{FORTRAN FILE}. The data from these files shall be read using
the \telfile{GET\_FREE\_ID} subroutine,
the \telfile{T3DBI1} logic unit for binary data file 1 and the \telfile{T3DBI2}
logic unit for binary data file 2,

\item One or two formatted data files, as specified by the keywords
\telkey{FORMATTED DATA FILE 1} and \telkey{FORMATTED DATA FILE 2}. These files
can be used to provide data to the program, the user has to handle their
reading within the \telkey{FORTRAN FILE}. The data from these files shall be
read using the \telfile{GET\_FREE\_ID} subroutine,
the \telfile{T3DFO1} logic unit for formatted data file 1 and the
\telfile{T3DFO2} logic unit for formatted data file 2,

\item A binary results file, as specified by the keyword \telkey{BINARY
RESULTS FILE}. This file can be used to store
additional results, the user has to handle their writing within the FORTRAN
program using the \telfile{GET\_FREE\_ID} subroutine
and the \telfile{T3DRBI} logic unit,

\item A formatted results file, as specified by the keyword \telkey{FORMATTED
RESULTS FILE}. This file can be used to store additional results (for instance,
results readable by an external 1D simulation code used for a coupling with
TELEMAC), the user has to handle their writing within the FORTRAN program using
the \telfile{GET\_FREE\_ID} subroutine and the \telfile{T3DRFO} logic unit.
\end{itemize}

The read or write operations from/into these files should be thoroughly managed
by the user (the files are opened and closed by the program). That management
can be performed from any point which the user can gain access to.
The logic unit numbers are stated in the \telfile{DECLARATIONS\_TELEMAC3D}
module and the user can access them through a
\telfile{USE DECLARATIONS\_TELEMAC3D} command at the beginning of a
subroutine. For instance, using a file for providing the initial conditions
will result in managing that file within the \telfile{CONDIM} subroutine
or the subroutines called by it (e.g.: \telfile{USER\_CONDI3D\_H} etc.).
Likewise, using a file for inserting boundary conditions will be possible at
the \telfile{USER\_BORD3D} subroutine.
In case of conflicting statements, one can use,
for example: \telfile{USE DECLARATIONS\_TELEMAC3D, ONLY: T3DBI1}.

\section{The dictionnary file}

This is a file containing all the information on the keywords (French name,
English name, default value, type, keywords documentation). This file can be
read by the user, but it should not be modified in any case.

\section{Topographic and bathymetric data}
\label{sec:topo}
Topographic and bathymetric data may be supplied to \telemac{3D} at three
levels:

\begin{itemize}
\item Either directly in the \telkey{GEOMETRY FILE} by a topographic
or bathymetric value associated with each mesh node.
In this case, the data are processed while the mesh is being built using
SALOME-HYDRO, Janet, Blue Kenue or MATISSE,
or when \stbtel is run before \telemac{3D} is started.
\stbtel reads the information in one or more bottom topography files (5 at most)
and interpolates at each point in the domain,

\item Or in the form of a cluster of points with elevations that have
no relation with the mesh nodes, during the \telemac{3D} computation.
\telemac{3D} then makes the interpolation directly with the same algorithm as
\stbtel.
The file name is provided by the keyword \telkey{BOTTOM TOPOGRAPHY FILE}.
Unlike \stbtel, \telemac{3D} only manages one single bottom topography file.
This may be in SINUSX format or more simply a file consisting of three columns
X, Y, Z.
The SINUSX format is described in the RUBENS user manual,

\item Or using the \telfile{USER\_T3D\_CORFON} subroutine.
This is usually what is done for schematic test cases.
\end{itemize}

In all cases, \telemac{3D} offers the possibility of smoothing the bottom
topography in order to obtain a more regular geometry.
The smoothing algorithm can be iterated several times depending on the degree of
smoothing required.
The keyword \telkey{NUMBER OF BOTTOM SMOOTHINGS} then defines the number of
iterations carried out in the \telfile{T3D\_CORFON} subroutine.
The default value of this keyword is 0.
