\chapter{General parameter definition for the computation}
\label{ch:gen:par:def:comp}
General setup of the computation is only done in the steering
file.

Time information is supplied by the three keywords:
\telkey{TIME STEP} (real set at 1. by default), \telkey{NUMBER OF TIME STEPS}
(integer set at 1 by default) and \telkey{DURATION} (real set at 0. by default).
The first one defines the time separating two consecutive instants of the
computation (but not necessarily two withdrawals from the results file).
The total duration of the computation may be supplied by means
of a number of time steps (keyword \telkey{NUMBER OF TIME STEPS})
or in the form of a total simulation period expressed in seconds
(keyword \telkey{DURATION}).
In the former case, the total duration is obviously equal to the time step value
multiplied by the number of time steps.

If a steering file contains both keywords \telkey{DURATION}
and \telkey{NUMBER OF TIME STEPS}, \telemac{2D} uses the one that produces
the longer simulation.
In addition, if the keyword \telkey{DURATION} is used and does not correspond
to a whole number of time steps, \telemac{2D} will take the integer immediately
higher.

By default the initial time is equal to 0.~s.
Since release v8p5, it can be changed by setting the keyword
\telkey{INITIAL TIME} (default = 0.~s).
It avoids to change previous hard-coded value 0. in \telfile{CONDIN} subroutine.
If input data is defined from a specific time and the user wants to start the
computation from another time, it is a way to do it without changing input data.

The date and hour corresponding to the initial state of the computation
are supplied by the keywords \telkey{ORIGINAL DATE OF TIME} (YYYY~;MM~;DD)
whose default value is (1900;1;1) i.e. January 1$^{\rm{st}}$ 1900
and \telkey{ORIGINAL HOUR OF TIME} (HH;MM;SS)
whose default value is (0;0;0) i.e. midnight.
This is particularly important if the tide generating forces are taken
into account (see \ref{sec:astral:pot}) and are generally necessary
when using tidal harmonic constituents databases.

The title of the computation is specified by the keyword \telkey{TITLE}.


\section{Criteria for stopping a computation}

Independently of normal time indications (number of time steps and time step
value), \telemac{2D} offers two possibilities for conditionally stopping the
computation:

\begin{itemize}
\item \underbar{Stopping when reaching a steady state}:
With this function, it is possible to start a computation, simulate a transient
flow and stop the computation when a steady state is reached.
The last time step in the results file created in this way can be used as an
initial state for other computations (e.g. tracer transport).
The test is triggered by indicating YES for the logical keyword
\telkey{STOP IF A STEADY STATE IS REACHED} (default = NO).
It is then possible to define the permissible area of tolerance
using the keyword \telkey{STOP CRITERIA}
(default value = (1.E-4 ; 1.E-4 ; 1.E-4)).
This keyword is an array of three real numbers, representing the tolerance
assigned to the velocity, water depth and tracer.
The computation is stopped when the absolute increment values of these variables
between two time steps at all nodes are below the limits indicated.
Assessing the right criterion depends on the case under study.
It should be stressed, however, that this function is inoperative
in the case of fundamentally non-stationary flows
such as Karman eddies behind bridge piers,

\item \underbar{Stopping in cases of divergence}:
This function is used to stop a computation if there is divergence.
The principle is the same as in the previous case.
The option is activated with the keyword \telkey{CONTROL OF LIMITS}
(default = NO).
The extreme values are indicated with the keyword \telkey{LIMIT VALUES}.
This is a table of 8 real numbers corresponding successively to:
\begin{itemize}
\item The minimum depth value for $H$ (by default -1,000~m),

\item The maximum depth value for $H$ (by default +9,000~m),

\item The minimum velocity value for $U$ (by default -1,000~m/s),

\item The maximum velocity value for $U$ (by default +1,000~m/s),

\item The minimum velocity value for $V$ (by default -1,000~m/s),

\item The maximum velocity value for $V$ (by default +1,000~m/s),

\item The minimum tracer value (by default -1,000),

\item The maximum tracer value (by default +1,000).
\end{itemize}
\end{itemize}

\section{Control sections}
\label{sec:contr:sect}
A control section offers the possibility of obtaining the instantaneous
and cumulated flow rates through a specific segment of the domain.

The weak formulation of the no-flux boundary condition through solid boundaries
raises a theoretical problem for computing the flow rates.
Either they are compatible with the results file, or they are compatible
with the weak formulation.
To be compatible with the weak formulation, use the keyword
\telkey{COMPATIBLE COMPUTATION OF FLUXES} (default = NO).
The difference may reach a few percents.

It is also possible to obtain the cumulated flow rates in the listing
for each control section
by activating the logical keyword \telkey{PRINTING CUMULATED FLOWRATES}
(default value = NO).
In that case, to improve the quality of results, the treatment of the control
section is done at each time step and not only at each time step concerned
by a printing on output listing.

The control sections can be managed using 2 different procedures.
The first one uses only a keyword and is not valid when running in parallel mode.
The second one (available since release 6.0) is based on an external
configuration file and is compatible with the parallel mode.
It is strongly recommended to use the new procedure.
The old procedure will be probably removed in a future release.


\subsection{Configuration with keywords only}

The section is defined using the keyword \telkey{CONTROL SECTIONS},
which is an array of pairs of integers separated by semi-colons,
containing the numbers of the beginning and the ending point of the section.

For example, the values: 611;54 ; 651;5210 define 2 control sections.
The first one is defined between points 611 and 54,
the second one between points 651 and 5210.

The results concerning the flow rates are written by \telemac{2D}
on the output control listing.
This information is the value of the instantaneous flow rate
and the cumulated positive and negative flow rates
(volume going through the section calculated from the beginning of the
simulation).
The sign is determined with the following rule:
going from the beginning to the ending point of the section,
the flow is positive when going from right to left.

The user may also use the subroutine \telfile{FLUXPR} (\bief library)
to exploit information connected with the control sections.


\subsection{Configuration with an external file}

The user must supply the name of the sections configuration file
using the keyword \telkey{SECTIONS INPUT FILE}.

In parallel mode, this file will be modified by the mesh partitioner
so that it corresponds locally to every sub-domain.

The file format is the following:

\begin{itemize}
\item one comment line (free but must be here),

\item two integers: number of sections and steering integer.
For the steering integer, the convention is as follow:
  \begin{itemize}
  \item if negative (< 0): node numbers (global numbers) are given,
  \item if equal to 0: coordinates are given.
Note that if positive (> 0) and parallel mode, coordinates can also given,
BUT the combo positive steering integer (> 0) and serial mode does not work!
  \end{itemize}

\item two lines per section:

\begin{itemize}

\item 24 characters for a section name, followed by:

\item begin and end node number or begin and end coordinates.

\end{itemize}
\end{itemize}

Example:
\begin{lstlisting}[language=bash]
# Control sections definition
5  -1
Wesxan_outflow
46 70
Wesxan_Middle
639 263
Wesxan_Inflow
480 414
Wesxan_crazy
142 147
Wesxan_even_worse
144 7864
\end{lstlisting}

Headers and printouts on control sections may be modified in
\telfile{FLUXPR\_TELEMAC2D} subroutine (\telemac{2D} library).

The printouts will be in the file named by the keyword
\telkey{SECTIONS OUTPUT FILE}.


\section{Computation of fluxes over lines (FLUXLINE)}

It is possible to compute fluxes over lines by using
\telkey{FLUXLINE} = YES (default = NO).
The input file with data on cross-sections has to be given
with the keyword \telkey{FLUXLINE INPUT FILE}.
The format of this ASCII file is:
\begin{itemize}
\item the first line contains the number of lines to be read $n$,
\item and then $n$ lines of 9 floating numbers describing each fluxline.
\end{itemize}
