\chapter{Recommendations}
\label{ch:reccom}
The purpose of this chapter is to provide the user with advices on using the
software.
\section{Mesh}

Some precautions need to be taken when building the mesh.
The following list should help, but it is not exhaustive of course.

\begin{itemize}
\item A liquid boundary should consist of at least 5 points, with 10 being
preferable,

\item In the case of a river mesh, and in particular for simulations of
low-flow periods, it is essential to refine the elements in the low-water bed
so as to ensure at least 3-4 points for conveying the flow.
If this rule is not followed, the results will be of poor quality.
In this case, it is possible to build the mesh of the low-water bed using
regular gridding available in most of mesh generators,

\item In domains with steep gradients in the topography or bathymetry,
the slope mesh must be refined if the current is not tangential to it,

\item It is preferable for triangles to be as nearly equilateral as possible,
as this type of element often gives the best results.
However, in the case of river meshes, it is sometimes interesting to elongate
the grid cells in the direction of the current, in order to reduce
the number of computation points and hence the simulation time.
\end{itemize}


\section{Initial conditions}

The technique most commonly used for maritime domains subject to tidal effects
is to initialize the free surface with a value corresponding to high tide
and the velocities with zero, and then gradually empty the domain.
Since the tidal solutions coming form Oregon State University (like TPXO global
solution), it has been possible to use them to initialize the computation.
It may initialize both water depth and horizontal velocity components.

In the case of river domains, two techniques are often used.
If the domain is relatively small (i.e. the bed level does not vary much
between upstream and downstream), the computation can be initialized
with constant elevations, by setting the value that will be prescribed
downstream of the computation domain as initial elevation.
Inflow is then gradually introduced from upstream.
This technique cannot be used if the model domain is very large,
as the initial elevation generally means that there will be a dry area upstream
of the model.
In this case, it is relatively easy, in the \telfile{USER\_CONDIN\_H} subroutine,
to initialize an elevation with a tilted plane
(the value of the elevation is proportional to the X or Y values)
and to introduce the nominal inflow progressively.
Another possibility is to use the free surface initialization implemented
in FUDAA-PREPRO.
This function offers the possibility to specify, in a very easy way, a free
surface slope defined by a longitudinal profile prescribed as a set of points.


\section{Numerical parameter definition}

\subsection{Type of advection}

Taking into account the recent improvements of \telemac{2D} in this field, the
following configuration can practically be considered as a ``quasi universal''
configuration (even in parallel mode):
\begin{lstlisting}[language=bash]
TYPE OF ADVECTION :  1 ; 5
\end{lstlisting}
Models with steep bottom topography gradients and tidal flats very often pose
serious difficulties (oscillations of the free surface, long computation times,
etc.).
In the light of experience, the configuration that appears to be best in such
cases is as follows:
\begin{lstlisting}[language=bash]
TREATMENT OF THE LINEAR SYSTEM = 2
FREE SURFACE GRADIENT COMPATIBILITY = 0.9
\end{lstlisting}

\subsection{Solver}

When using primitive equations (which is no longer recommended), the solver
giving the best results in terms of computation time is GMRES (keyword value 7).
In this case, it is sometimes useful to configure the dimension of the Krylov
space in order to optimize computation time.
The larger the dimension, the more time is required to run an iteration,
but the faster the system converges.
The user is therefore strongly advised to run simulations over a few time steps
by varying the keyword \telkey{SOLVER OPTION}
(and \telkey{OPTION FOR THE SOLVER}) so as to reach the best compromise between
computation time for one iteration and the number of iterations, remembering
that the more points there are in the mesh the higher the optimum value.
This optimum value generally varies from 2 (small meshes) to 4 or 5 (large
meshes).
When using this solver, the optimum value for the time step (in terms of
computational time) is generally reached when the convergence occurs with 10 to
20 iterations.

When using the wave equation, the recommended solver is the conjugate gradient
(value 1).
In that case, the optimum value for the time step is generally reached
when the convergence occurs with 30 to 50 iterations.


\section{Special types of programming}


\subsection{Changing bottom topography between two computations}

The \telfile{CORFON} subroutine is used to change the bottom topography read
from the \telkey{GEOMETRY FILE}.
Everything is programmed so that this change is only done once.
The list of operations is as follows:
\begin{itemize}
\item Reading of the geometry,

\item Bottom correction with \telfile{USER\_CORFON} subroutine.
\end{itemize}

If a computation is being continued, the bottom from the previous computation
results file is used, if there is one.
Any change of the \telfile{USER\_CORFON} subroutine for a continued computation
will therefore be inoperative if the bottom topography is saved in the results
file, even if \telfile{USER\_CORFON} is actually called.

The procedure for changing bottom topography between two successive computations
is as follows:
\begin{itemize}
\item Run an initial computation without saving the bottom topography or water
depth, but saving the free surface,

\item Modify the \telfile{USER\_CORFON} subroutine,

\item Continue the computation.
\telemac{2D} will then use the new bottom topography and as it only finds the
free surface in the results of the previous computation, it will recalculate the
new water depth as being the old free surface minus the new bottom topography.

\end{itemize}


\section{Tidal flats}

The following explanations concern the Finite Elements option.
In Finite Volumes options (see keyword \telkey{EQUATIONS}), mass-conservation is
ensured on tidal flats and the depth remains positive.
However, e.g. in the case of the Malpasset dam break test-case, these explicit
techniques will be much more time-consuming (factor around 10).

The treatment of tidal flats is a very strategic issue in flood and dam-break
flood wave computations.
Over the years a number of specific procedures have been developed in
\telemac{2D} to cope with this difficulty.
Historically, the basic option \telkey{TREATMENT OF THE TIDAL FLATS} = 2
consisted in removing from the computation the dry elements.
This option cannot be used in parallel computations.
With this option, the keyword \telkey{MINIMUM VALUE OF DEPTH} is used to decide
whether an element is dry or not.
This option is not generally recommended, but proved to be more stable with
quasi-steady flows in rivers.

The preferred option is obtained with \telkey{TREATMENT OF THE TIDAL FLATS} = 1.
In this case, all the finite elements are kept in the computation,
which implies a specific treatment of dry points, especially when divisions
by the depth occur in the equations.
For example the friction terms as they appear in the non-conservative momentum
equations would be infinite on dry land, and are limited in the computation.
Mass-conservation is guaranteed with this option, but it is never imposed
that the depth should remain positive, and slightly negative depths may appear
(any correction with the keyword \telkey{H CLIPPING} would spoil the
mass-conservation).

The option \telkey{TREATMENT OF THE TIDAL FLATS} = 3 is basically the same as
option 1, but on partially dry elements, a porosity coefficient is applied to
take into account the fact that in reality, the finite element has a size
limited to its wet part.
This option has been designed mainly for dam break studies, though users report
a good behavior in quasi-steady flows.
Unless specific reasons and waiting for more convincing tests, option 1 is
recommended rather than 3.

When using option 1 or 3, it is possible to use a specific treatment concerning
the negative depths by selecting the appropriate value for the keyword
\telkey{TREATMENT OF NEGATIVE DEPTHS}.
The possibilities are:

\begin{itemize}
\item 0: no treatment. The negative depths are left unchanged,

\item 1: smoothing of negative depth (default value),

\item 2: ''Flux control'', by segment,

\item 3: ''Flux control'' ERIA, by triangular element.
\end{itemize}
The last two treatments mean that some fluxes between points or elements
may be limited to avoid negative depths.
If using options 2 or 3 with tidal flats, it is mandatory to set the following
keywords:
\begin{itemize}
\item \telkey{MASS-LUMPING ON H} = 1.,
\item \telkey{CONTINUITY CORRECTION} = YES,
\item \telkey{SUPG OPTION} for water depth = 0 (no SUPG upwinding on depth).
\end{itemize}

When using option 1, it is possible to set the limit value for the smoothing
using the keyword \telkey{THRESHOLD FOR NEGATIVE DEPTHS}
which default value is 0.\\

Hereafter are general recommendations when there are tidal flats in your domain:

\begin{itemize}
\item of course, use the keyword \telkey{TIDAL FLATS} = YES,

\item avoid tidal flats every time it is possible,
e.g. very steep banks can sometimes be replaced by a vertical wall,

\item refine the mesh on dykes or other features that will be submerged
and that have a critical effect on flooding. Preferably use the wave equation.
\end{itemize}

Here are the main options chosen for a quasi-steady flow
(Wesel-Xanten case originally provided by BAW):
\begin{lstlisting}[language=bash]
VELOCITY PROFILES                       = 4;0
TURBULENCE MODEL                        = 1
VELOCITY DIFFUSIVITY                    = 2.
TIDAL FLATS                             = YES
OPTION FOR THE TREATMENT OF TIDAL FLATS = 1
TREATMENT OF NEGATIVE DEPTHS            = 2
FREE SURFACE GRADIENT COMPATIBILITY     = 0.9
H CLIPPING                              = NO
TYPE OF ADVECTION                       = 1;5
SUPG OPTION                             = 0;0
TREATMENT OF THE LINEAR SYSTEM          = 2
SOLVER                                  = 2
PRECONDITIONING                         = 2
SOLVER ACCURACY                         = 1.E-5
CONTINUITY CORRECTION                   = YES
\end{lstlisting}
The wave equation (\telkey{TREATMENT OF THE LINEAR SYSTEM} = 2) proved here
to be more stable than primitive equations.
These options are also convenient for the Malpasset dam-break computation,
and can thus be taken as a starting point for a new case.

The keyword \telkey{OPTION FOR THE DIFFUSION OF VELOCITIES} should normally be
set to 2, as it is the correct theoretical formula, however the simplified form
corresponding to option 1 is preferred, because it avoids the problem of
division by 0 on dry zones.
So far no clear test-case proved the superiority of option 2.
