\chapter{Parallelism}
\label{ch:paral}
\telemac{2D} is generally run on single-processor computers of the workstation
type.
When simulations call for high-capacity computers and in the absence of a
super-computer, it may be useful to run the computations on multi-processor
(or multi-core) computers or clusters of workstations.
A parallel version of \telemac{2D} is available for use with this type of
computer architecture.

The parallel version of \telemac{2D} uses the MPI library, which must therefore
be installed beforehand.
The interface between \telemac{2D} and the MPI library is the parallel library
common to all modules of the TELEMAC system (in folder /sources/utils/parallel).

Information on the use of the parallel version is given in the system
installation documents.

Initially, the user must specify the number of processors used by means of the
keyword \telkey{PARALLEL PROCESSORS}.
The keyword may have the following values:

\begin{itemize}
\item 0: Use of the classical version of \telemac{2D}
(in a parallel version same as 1),

\item 1: Use of the parallel version of \telemac{2D} with one processor,

\item N: Use of the parallel version of \telemac{2D} with the specified number
of processors, here N.
\end{itemize}

%Domain decomposition and results file combination operations are now automatic
%and handled completely by the start-up procedure.
The parallel version uses a domain decomposition method to split/merge the mesh
for each processor.
Those extra steps are done using PARTEL/GRETEL and are launched by the Python
execution scripts.

Note that the number of processors/cores can be given as an argument for the
launching command and not as a hard-coded keyword in the \telkey{STEERING FILE}
(e.g. telemac2d.py --ncsize=4 cas.txt will run \telemac{2D} on 4 processors).
