\chapter{Secondary currents}
\label{ch:sec:curr}
In a curved channel, the flow experiences a radial acceleration and centrifugal
forces act in proportion to the mean velocity.
In turn, the surface of the water is tilted radially on the outer bank
to produce a super-elevation sufficient to create a pressure gradient 
to balance the average centrifugal force.
At shallower depths, the centrifugal force exceeds the pressure force,
whence the resultant force drives the fluid outwards.

\telemac{2D} allows to take into account the effect of these secondary currents.
To activate this, key-word \telkey{SECONDARY CURRENTS} must be set to TRUE
(default = FALSE).
User can also manage some coefficients used in the resolved equation
(see release note 7.0 for theoretical aspects and for more details).
For instance, the production term in the advection-diffusion equation linearly
depends on a coefficient $A_s$ which can be calibrated using the keyword
\telkey{PRODUCTION COEFFICIENT FOR SECONDARY CURRENTS} (default = 7.071).
In the same way, the dissipation part can be modified by varying
the coefficient $D_{s}$ using the keyword
\telkey{DISSIPATION COEFFICIENT FOR SECONDARY CURRENTS} (default = 0.5).
Further details can be found in~\cite{wang2014secondary}.
\\

If there are open boundaries, the user cannot set any value to the keyword
\telkey{PRESCRIBED TRACERS VALUES} for the streamwise vorticity $\Omega$,
the variable which is considered as a tracer by \telemac{2D} to model the
secondary currents.
Indeed, they are already hard-coded to be 0.\\

An example of use of the secondary currents is given in the validation case
seccurrents (in the folder examples/telemac2d). 
This example is described and well documented in the validation manual of
\telemac{2D}.
