\chapter{BIOMASS Module}

The BIOMASS module is a water quality module which allows the calculation of algal biomass.
It estimates the extent of vegetal colonization in terms of various parameters:
sunlight, water temperature, fertilization degree, water renewal ratio,
water turbidity and toxicity \cite{gosse_biomass_1983}.
The BIOMASS module takes into account five tracers:

\begin{itemize}
\item phytoplankton biomass PHY,
\item the principal nutrients influencing its production (phosphorus, nitrogen)
  as well as the associated mineral forms, namely:

\begin{itemize}
\item dissolved mineral phosphorus assimilable by phytoplankton PO$_4$,
\item degradable phosphorus not assimilable by phytoplankton POR,
\item dissolved mineral nitrogen assimilable by phytoplankton NO$_3$,
\item degradable nitrogen not assimilable by phytoplankton NOR.
\end{itemize}
\end{itemize}

These variables are all expressed in mg/l except biomass that is expressed in $\mu$g(Chlorophyl a)/l.\\

We assume these substances acting as tracers,
i.e. they are carried and dispersed in the water mass.
In addition, they react with each other through biochemical processes.\\

The following sections explain the internal source terms.

\section{Phytoplankton}

\subsection{Algal growth}

The algal growth rate $CP$ (d$^{-1}$) is given by:

\begin{equation}
  CP = C_{max} RAY g_1 LNUT \alpha_1,
\end{equation}

with $C_{max}$ = algal growth maximum rate at 20$^{\circ}$C; one can take $C_{max}$ = 2.
$RAY$ represents the effect of sunlight on algal growth;
this dimensionless parameter ranges between 0 and 1;
$g_1$ represents the effect of temperature on algal growth;
$g_1 = T/20$, where $T$ is the water temperature ($^{\circ}$C) (valid for 5$^{\circ}$C < $T$ < 25$^{\circ}$C).
$LNUT$ represents the effects of phosphoric and nitric nutrients on algal growth.
$\alpha_1$ = water toxicity coefficient for algae ($\alpha_1$ = 1 in the absence of toxicity).\\

In 2D, $RAY$ is calculated by the Smith formula averaged over the vertical:

\begin{equation}
  RAY = \frac{1}{k_e h} \log \left( \frac{I_0 + \sqrt{IK^2+I_0^2} }{ I_h + \sqrt{IK^2+I_h^2} }  \right),
\end{equation}

where $k_e$ is the extinction coefficient of solar rays in water (m$^{-1}$).
It is calculated either by the Secchi depth $Z_s$ using the Atkins formula: $k_e$ = 1.7/$Z_s$, or,
if $Z_s$ is unknown, by the Moss relation: $k_e$ = $k_{pe}$+$ \beta $ [PHY],
where $k_{pe}$ is the coefficient of vegetal turbidity without phytoplankton (m$^{-1}$)
and $ \beta $  the Moss coefficient ($ \beta \approx$ 0.015).
$IK$ is a calibrating parameter (W/m$^2$), of an order of magnitude 100.
$I_0$ is the flux density of solar radiation on the surface (W/m$^2$)
and $I_h$ is the flux density of solar radiation at the bed bottom (W/m$^2$), calculated as:

\begin{equation}
  I_h = I_0 \exp (-k_e h).
\end{equation}

In 3D, $RAY$ is calculated by a generalisation of the Smith formula
over the vertical:

\begin{equation}
  RAY = \frac{1}{k_e \Delta z} \log \left( \frac{I_0 + \sqrt{IK^2+I_0^2} }{ I_{\Delta z} + \sqrt{IK^2+I_{\Delta z}^2} }  \right),
\end{equation}

with everything like in 2D except $\Delta z$ the height of the prism in 3D and
$I_{\Delta z}$ is the flux density of solar radiation at the considered depth
(W/m$^2$), calculated as:

\begin{equation}
  I_{\Delta z} = I_0 \exp (-k_e \Delta z),
\end{equation}


$LNUT$ is calculated by the formula:

\begin{equation}
  LNUT = \min \left( \frac{[PO_4]}{KP+[PO_4]}, \frac{[NO_3]}{KN+[NO_3]} \right),
\end{equation}

with $KP$ = phosphate half-saturation constant (mg/l) (about 0.005 mgP/l),
and $KN$ = nitrate half-saturation constant (mg/l) (about 0.03 mgN/l).\\

Note: Nutrients affect phytoplankton growth PHY only by limiting the factor $LNUT$.
When [PO$_4$] and [NO$_3$] are high enough, $LNUT$ is close to 1 and
phytoplankton evolution no longer depends on nutrients.
In this case, there is no need to model the cycles of phosphorus and nitrogen
for simulating the evolution of phytoplankton.

\subsection{Algal disappearance}

The algal disappearance rate $DP$ (d$^{-1}$) is given as:

\begin{equation}
  DP = (RP+MP) g_2,
\end{equation}

with $RP$ = algal biomass respiration rate at 20$^{\circ}$C (d$^{-1}$),
$MP$ = algal biomass disappearance rate at 20$^{\circ}$C (d$^{-1}$).
$g_2$ represents the effect of temperature on algal disappearance.
$g_2 = T/20$ (valid for 5$^{\circ}$C < $T$ < 25$^{\circ}$C).
$MP$ is given by the following relation:

\begin{equation}
  MP = M_1 + M_2 [PHY] + \alpha_2,
\end{equation}

with $M_1$ and $M_2$ = algal mortality coefficients at 20$^{\circ}$C,
$\alpha_2$ = water toxicity coefficient for algae.

\section{Nitric and phosphoric nutrients}

The following physical and biochemical parameters are used
to describe the processes influencing the evolution of nitric and phosphoric nutrients:

\begin{itemize}
\item $fp$: average proportion of phosphorus in the cells of living phytoplankton (mgP/$\mu$gChlA),
\item $dtp$: proportion of directly assimilable phosphorus in dead phytoplankton ($\%$),
\item $k_1$: transformation rate of POR into PO$_4$ through bacterial mineralization (d$^{-1}$),
\item $k_2$: transformation rate of NOR into NO$_3$ through heterotrophic
  and autotrophic bacterial mineralization (d$^{-1}$),
\item $fn$: average proportion of directly assimilable nitrogen in living phytoplankton (mgN/$\mu$gChlA),
\item $dtn$: proportion of directly assimilable nitrogen in dead phytoplankton ($\%$),
\item $F_{POR}$: deposition flux of non-algal organic phosphorus (g/m$^2$/s).
  $F_{POR} = W_{POR} [POR]$, $W_{POR}$ is the sedimentation velocity of non-algal organic phosphorus (m/s),
\item $F_{NOR}$: deposition flux of non-algal organic nitrogen (g/m$^2$/s).
  $F_{NOR} = W_{NOR} [NOR]$, $W_{NOR}$ is the sedimentation velocity of non-algal organic nitrogen (m/s).
\end{itemize}

Note: sediment transport and resulting bed changes are not modeled in the BIOMASS model.
Deposition is only represented by the deposition flux and,
once organic matter is deposited,
it no longer appears in the equations and can no longer be resuspended.
These deposition fluxes therefore correspond to a definitive loss of mass.

\section{Solved equations in 2D}

The BIOMASS model equations are described below, detailing the internal source terms $S_{intern\_i}$.\\

Tracer $\#$1: phytoplankton biomass

\begin{equation}
  F([PHY]) = (CP-DP) [PHY].
\end{equation}

Tracer $\#$2: assimilable mineral phosphorus

\begin{equation}
  F([PO_4]) = fp(dtp DP - CP) [PHY] + k_1 g_2 [POR].
\end{equation}

Tracer $\#$3: non-assimilable phosphorus

\begin{equation}
  F([POR]) = fp(1-dtp) DP [PHY] - k_1 g_2 [POR] - \frac{F_{POR}}{h}.
\end{equation}

Tracer $\#$4: assimilable mineral nitrogen

\begin{equation}
  F([NO_3]) = fn (dtn DP - CP) [PHY] + k_2 g_2 [NOR].
\end{equation}

Tracer $\#$5: non-assimilable nitrogen

\begin{equation}
  F([NOR]) = fn (1- dtn) DP [PHY] - k_2 g_2 [NOR] - \frac{F_{NOR}}{h},
\end{equation}

with $C_1$ = [PHY], $C_2$ = [PO$_4$], $C_3$ = [POR], $C_4$ = [NO$_3$] and $C_5$ = [NOR],
the matrices (5 $\times$ 5) $[\lambda]$ and $[\mu]$
%containing the coefficients $\lambda_i^j$ and $\mu_i^j$
are written as (only non-zero terms are included):

\begin{equation}
\lambda_i^j = \frac{1}{86400}
\left(
%  \begin{array}{c;{2pt/2pt}c;{2pt/2pt}c;{2pt/2pt}c;{2pt/2pt}c}
  \begin{array}{ccccc}
    CP-DP & 0 & 0 & 0 & 0 \\ %\hdashline[2pt/2pt]
    fp (dtp DP -CP) & 0 &  k_1 g_2 & 0 & 0 \\ %\hdashline[2pt/2pt]
    fp (1-dtp) DP   & 0 & -k_1 g_2 & 0 & 0 \\ %\hdashline[2pt/2pt]
    fn (dtn DP -CP) & 0 &        0 & 0 &  k_2 g_2 \\ %\hdashline[2pt/2pt]
    fn (1-dtn) DP   & 0 &        0 & 0 & -k_2 g_2
  \end{array}
\right)
\end{equation}

$$
  \mu_i^j =
  \begin{pmatrix}
   0 & 0 & 0 & 0 &  0 \\
   0 & 0 & 0 & 0 &  0 \\
   0 & 0 & -W_{POR} & 0 & 0 \\
   0 & 0 & 0 & 0 &  0 \\
   0 & 0 & 0 & 0 & -W_{NOR}
  \end{pmatrix}
$$

The terms $\lambda_i^0$ and $\mu_i^0$ are zero for every $i$.\\

Divisions by 86,400 are performed to scale down time to one second.

\section{Solved equations in 3D}

The BIOMASS model equations are described below, detailing the internal source terms $S_{intern\_i}$.\\

Tracer $\#$1: phytoplankton biomass

\begin{equation}
  F([PHY]) = (CP-DP) [PHY].
\end{equation}

Tracer $\#$2: assimilable mineral phosphorus

\begin{equation}
  F([PO_4]) = fp(dtp DP - CP) [PHY] + k_1 g_2 [POR].
\end{equation}

Tracer $\#$3: non-assimilable phosphorus

\begin{equation}
%  F([POR]) = fp(1-dtp) DP [PHY] - k_1 g_2 [POR] - F_{POR}.
  F([POR]) = fp(1-dtp) DP [PHY] - k_1 g_2 [POR].
\end{equation}

Tracer $\#$4: assimilable mineral nitrogen

\begin{equation}
  F([NO_3]) = fn (dtn DP - CP) [PHY] + k_2 g_2 [NOR].
\end{equation}

Tracer $\#$5: non-assimilable nitrogen

\begin{equation}
%  F([NOR]) = fn (1- dtn) DP [PHY] - k_2 g_2 [NOR] - F_{NOR},
  F([NOR]) = fn (1- dtn) DP [PHY] - k_2 g_2 [NOR],
\end{equation}

with $C_1$ = [PHY], $C_2$ = [PO$_4$], $C_3$ = [POR], $C_4$ = [NO$_3$] and $C_5$ = [NOR],
the matrices (5 $\times$ 5) $[\lambda]$ and $[\mu]$
%containing the coefficients $\lambda_i^j$ and $\mu_i^j$
are written as (only non-zero terms are included):

\begin{equation}
\lambda_i^j = \frac{1}{86400}
\left(
%  \begin{array}{c;{2pt/2pt}c;{2pt/2pt}c;{2pt/2pt}c;{2pt/2pt}c}
  \begin{array}{ccccc}
    CP-DP & 0 & 0 & 0 & 0 \\ %\hdashline[2pt/2pt]
    fp (dtp DP -CP) & 0 &  k_1 g_2 & 0 & 0 \\ %\hdashline[2pt/2pt]
%    fp (1-dtp) DP   & 0 & -k_1 g_2 -W_{POR} & 0 & 0 \\ %\hdashline[2pt/2pt]
    fp (1-dtp) DP   & 0 & -k_1 g_2 & 0 & 0 \\ %\hdashline[2pt/2pt]
    fn (dtn DP -CP) & 0 &        0 & 0 &  k_2 g_2 \\ %\hdashline[2pt/2pt]
%    fn (1-dtn) DP   & 0 &        0 & 0 & -k_2 g_2 - W_{NOR}
    fn (1-dtn) DP   & 0 &        0 & 0 & -k_2 g_2
  \end{array}
\right)
\end{equation}

$$
  \mu_i^j =
  \begin{pmatrix}
   0 & 0 & 0 & 0 & 0 \\
   0 & 0 & 0 & 0 & 0 \\
   0 & 0 & 0 & 0 & 0 \\
   0 & 0 & 0 & 0 & 0 \\
   0 & 0 & 0 & 0 & 0
  \end{pmatrix}
$$

The terms $\lambda_i^0$ and $\mu_i^0$ are zero for every $i$.\\

Divisions by 86,400 are performed to scale down time to one second.

\begin{WarningBlock}{Warning:}
Since release 8.5 and as long as no good solution to implement the treatment of
sedimentation velocities (N and P) is found, this treatment has been commented
in the source code so that it is not possible to take them into account
\emph{in 3D}.
\end{WarningBlock}
