\chapter{\TelApy{} description}\label{ch:TelApy_description}

As mentioned in the introduction part (Section~\ref{ch:intro}), the \TelApy{}
module is used to control the APIs of \telemacsystem{} in the \python{}
programming language. The \telemacsystem{} APIs are developed in \fortran{}.
However, it is relatively easy to use these \fortran{} routines directly in
\python{} using the ``f2py'' tool from the \python{} NumPy
library~\cite{Peterson2009}. This tool will make it possible to compile
\fortran{} code to make it accessible and usable in \python{}. This compilation
step is directly integrated into the compilation of the \telemacsystem{} and is
thus transparent to the user. Moreover, in order to make the \TelApy{} module
more user friendly, a \python{} wrapper has been developed in order to
encapsulate and simplify the different API \python{} calls. This set of
transformation constitutes the \TelApy{} module.

The first section of this chapter is dedicated to the \fortran{} API of
\telemacsystem{}. Then, the \python{} \TelApy{} module is presented.

\section{\telemacsystem{} \fortran{} API description}
%------------------------------------------------------------------------------
%
An API (Application Programming Interface) is a library allowing to control the
execution of a program. Here is part of the definition from Wikipedia:\\

"In computer programming, an application programming interface (API) specifies
a software component in terms of its operations, their inputs and outputs and
underlying types. Its main purpose is to define a set of functionalities that
are independent of their respective implementation, allowing both definition
and implementation to vary without compromising each other.

In addition to accessing databases or computer hardware, such as hard disk
drives or video cards, an API can be used to ease the work of programming
graphical user interface components, to allow integration of new features into
existing applications (a so-called "plug-in API"), or to share data between
otherwise distinct applications. In practice, many times an API comes in the
form of a library that includes specifications for routines, data structures,
object classes, and variables." \\

The API’s main goal is to have control on a simulation while running a case.
For example, it must allow the user to stop the simulation at any time step,
retrieve some variables values and change them. In order to make this possible,
a \fortran{} structure called instance is used in the API\@. This informatic
structure is described in the paragraph~\ref{subsec:instan}. The instance
structure gives direct access to the physical memory of variables, and allows
therefore the variable control (see paragraph~\ref{subsec:var}). Furthermore,
based on modifications in \telemacsystem{} main subroutines, it is possible to
run hydraulic case time step by time step. This will be
presented in the paragraph~\ref{subsec:exec}. And finaly, the parallelism is
evoked (paragraph~\ref{subsec:para}).
All \fortran{} routines are available in the directory ``api'' of the
\telemacsystem{} repository.
sources.

Detailed information on each function can be find in the Doxygen documentation
to open it run the following command:

\begin{lstlisting}
firefox <root>/documentation/doxydocs/html/index.html
\end{lstlisting}

Replacing <root> by the path to your installation and firefox by your internet
browser.

If you are on the main branch, run the command below to generate it.

\begin{lstlisting}
doc_telemac.py -M doxydocs
\end{lstlisting}

\begin{WarningBlock}{Warning:}
\centering
 In the following sections, all presented API routines are related to
 \telemac{2D}. However, the API implementation of \telemacsystem{} modules is
 generic that is to say based on the same structure (in the following routines,
 the sequence "t2d" related to \telemac{2D} module can be replaced e.g. with
 "t3d" to use the \telemac{3D} module).
\end{WarningBlock}

\subsection{Structure}

The functions are dispatched in \fortran{} files as follow:
\begin{itemize}
  \item \verb!api_interface.f!: Contains the main function for all the api.
  \item \verb!api_handle_error.f!: Contains the handling of error messages.
  \item \verb!api_handle_var_t2d.f!: Contains all the get/set.
  \item \verb!api_instance_t2d.f!: Contains the function for the instance
    described in Section~\ref{subsec:instan}.
  \item \verb!api_run_t2d.f!: Contains the function for the execution of the
    computation described in Section~\ref{subsec:exec}.
\end{itemize}

\subsection{Instantiation}\label{subsec:instan}
An instance is a memory structure that gathers all the variables alterable
by the API\@. The definition of the ``instance'' structure is made in a
\fortran{} type dedicated to this purpose and is composed of:

\begin{itemize}
\item An index defining the instance ID\@.
\item A string which can contain error messages.
\item Some pointers to the concerned module variables. This is what makes it
  possible to manipulate the variables of the module by having a direct access
    to their memory location. The list of variables that can be accessed is
    given in the Tables below. Not all
    variables within the module are there. However adding a new variable is
    pretty straight forward, the procedure is described in
    Section~\ref{var_add}.
\end{itemize}

%Automatically generated by python list of variables
\input{latex/var_list_t2d.tex}
\input{latex/var_list_t3d.tex}
\input{latex/var_list_art.tex}
\input{latex/var_list_wac.tex}

In addition to the instance definition, the module includes all routines needed
to manipulate it (creation, deletion, and so on).

\subsection{Variable control}
\label{subsec:var}
The way in which the instance is defined (pointers) allows manipulation of
variables during the simulation. So, to get information on the variables the
following set of functions has been implemented:

\begin{itemize}

\item \verb!get_var_list! get the list of variables reachable with the API:

\item \verb!get_var_type! get the type of a variable.

\item \verb!get_var_size! get the size of a variable.

\item \verb!get_*!, \verb!set_*! access to a given index of a
  variable.

\item \verb!get_double_array! and \verb!set_double_array! to optimize
  access to a whole array instead of just a value a new pair of
functions was created.

\end{itemize}

The \verb!get_*! exist for 4 kind of types (boolean, integer, double/float,
string). However the type distinction is removed in \TelApy{} based on
\python{} benefits.

All of those functions are in \verb!api_interface.f! they all take as arguments
a short name (t2d, t2d, sis\ldots) and an id.

The detailed information for all the routines are available in the Doxygen
documentation.

\subsection{Computation control}\label{subsec:exec}

The computation control is carried out using some specific routines to launch
the simulation. These routines constitute a decomposition of the main program
of each \telemacsystem{} modules corresponding to the following different
computation steps:

\begin{itemize}
\item \verb!run_set_config_t2d!: Configuration setup. This function initialises
  the instance and the listing output. The instance, characterised by the
  \textbf{id} integer parameter, represents a run of \telemac{2D}.

\begin{CommentBlock}{Comment:}
\centering
  In the current version you can have only one instance of each module running
  at the same time.
\end{CommentBlock}

\item \verb!run_read_case_t2d!: Reading the \telemac{2D} steering file. This
  function reads the case file and set the variable of the \telemac{2D}
    steering file accordingly.

\begin{WarningBlock}{Warning:}
\centering
  With the API we are not using the temporary folder (this folder was created
  by the \python{} environment and all the file declared in the steering file
  where copied and renamed inside that folder) which means that the name and
  path given in the steering file will be used. This also this means that the
  same file cannot be used for multiple keywords.
\end{WarningBlock}

\item \verb!run_allocation!: Memory allocation. This function runs the
  allocation of all the data needed in \telemac{2D}. Any modifications to
    quantities of \telemac{2D} should be done before the call to that function.

\item \verb!run_init_t2d!: Initialization. This function will do the setting of
  the initial conditions of \telemac{2D}. It corresponds to the time-step 0 of
    a \telemac{2D} run.

\item \verb!run_timestep_t2d!: Computation function that runs one time-step
  of \telemac{2D} and writing the results. To compute all time steps, a
    loop on this function must be done.

\item \verb!run_timestep_compute_t2d!: Computation function that runs one
  time-step of \telemac{2D} without writing the results (only in \telemac{2D}).

\item \verb!run_timestep_res_t2d!: Computation function that runs the writing
  part of a time step on listing and/or files (only in \telemac{2D}).

\item \verb!run_finalize_t2d!: Finalization. This function concludes the run of
  \telemac{2D} and will deallocate all the arrays and delete the instance. To
    start a new execution of \telemac{2D} the function run\_set\_config must be
    run again.

\end{itemize}

For each routine defined above, the first argument is the identity number it is
allowing all computation variables to be linked with the corresponding instance
pointers. These routines are then called in the same order to insure a correct
execution of the computation in the API main program.

\subsection{Parallelisation}
\label{subsec:para}
All steps associated with parallel computation must be performed by the user
when he chooses to launch his calculation on several processors. In this case,
after initializing the MPI environment, the user must partition the input files
(geometry file, boundary conditions files, and so on) using the \fortran{}
function "partel". Then, when the calculation is complete, it is necessary to
merge each subdomains result files using the "gretel" routine of the
\telemacsystem{}. The MPI environment can then be closed.

All the get/set functions are using local numbering.
To handle global number two tools are available:
\begin{itemize}
  \item the function GLOBAL\_TO\_LOCAL\_POINT returns the local numbering of a
    global id
(0 if it is not on the partition).
  \item the array KNOLG returns the global number of a local point.
\end{itemize}

\section{\TelApy{} \python{} module}

It is relatively easy to use the \fortran{} API routines directly in \python{}
using the ``f2py'' tool of the \python{} Numpy library. This tool will make it
possible to compile \fortran{} code such as it is accessible and usable in
\python{}. For more details on this tool, the interested reader can refer
directly to \cite{Peterson2009}. However, using the advantage of the \python{}
language, it is possible to implement a wrapper in order to provide user
friendly function of the \fortran{} API\@. Thus, a \python{} overlay was
developed in order to encapsulate and simplify the different API \python{}
calls. The different \python{} functions written to simplify the use of API are
available in the directory "$HOMETEL/scripts/python3/telapy/$".

A \doxygen documentation (In \verb!<root>/documentation/doxypydocs! if you are
on the main branch, run \verb!doc_telemac.py -M doxypydocs! to generate it) is
available and allows the user to visualize \python{} classes, functions that
can be used as well as its input and output variables and so on.

In order to launch the \doxygen documentation, the user needs to copy and paste
the link \verb!$HOMETEL/documentation/doxypydocs/html/index.html! into his
favorite internet browser. Then, the user can navigate in the \doxygen
environment in order to find information.
